---
subtitle: "DATASCI 306: Lecture 17"
title: "Factors"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(nycflights13)
```

## Strings and Factors

R has two data types for representing categorical data types:

* Strings: character data
* Factors: integers representing categories with look up table for labels, can be ordered 

```{r}
example_str <- c("you", "say", "goodbye", "i", "say", "hello", "hello", "hello")
example_str
example_fct <- factor(example_str)
example_fct
```



Factors are used for:

* categorical variables (variables that have a fixed and known set of possible values)
* also useful when you want to display character vectors in a non-alphabetical order.

In R we typically represent categorical variables using factors.

## Strings vs. factors

Many times two types have the same function, why use one or the other?

- Want to order lexicographically: strings
- Want to order in other way: factors
- Treat as text (search, combine): strings
- Avoid errors from treating as text: factors

There used to be a space saving advantage to factors but that has been eliminated with newer versions of R.

## `forcats` package

`tidyverse` contains a package that has some tools for working with factors. Sometimes it is not automatically loaded by the tidyverse metapackage, in which case you must load it manually:

`library(forcats)`

`forcats` commands are prefixed by fct_ (compare stringr).


## Why factors?
To learn the basics, we'll start with a simple example

```{r}
birth_months = c("Jan", "Feb", "Sep", "Sep", "Dec", "Jan", "Jul", "Aug")  # categorical variable
```

The `possible` values of a categorical variable are called the `levels`. The levels of birth_months are `Jan, Feb, ..., Dec`. The actual values of birth_months are just called the `values`.

To create a factor, we must specify the levels and the values:

```{r}
month.abb
```

```{r}
birth_months_fac = factor(birth_months, levels=month.abb) |> print()
```

## Significance of levels

If you don't specify the levels, R will infer them based on all the observed values of the variable:

```{r}
factor(birth_months) |> str()
```

It's best to be explicit about the factor levels. This way, if you there are typos or data entry errors, you will catch them more easily

```{r}
bad = c("Jan", "Feb", "Sep", "Ser", "Dec", "Jan", "Jul", "Aug")
#                               ^ note typo
# factor(bad)
factor(bad, levels=month.abb)
```


## Order

One advantage of factors is that they can be ordered. This enables them to sort and plot in the way you would expect. 

Compare:
```{r}
sort(birth_months)
```

```{r}
sort(birth_months_fac)
```

If you specify the factor levels using the levels= option, then that will specify the default order. If you do not specify the levels, then they will be sorted alphabetically by default.

## levels function
You can just get the levels of a factor using the levels function

```{r}
levels(birth_months_fac)
```

```{r}
levels(factor(birth_months))
```

## General Social Survey - gss dataset
This is a sample of data from the [General Social Survey](https://gss.norc.org/), a long-running US survey conducted by the independent research organization NORC at the University of Chicago. This is available with tidyverse package.

We will use an augmented version of the gss data provided by the package
```{r}
gss18 <- read_csv('data/gss18.csv')
gss18 |> glimpse()
```

We will use this data to illustrate some common challenges you’ll encounter when working with factors.

You can't see the unique categorical values easily in a tibble. One way to view them is with count():

```{r}
gss18 |> count(race)
```

## Working with factors
The two main aspects when working with factors:

* changing the order of the levels
* changing the values of the levels

We will look at how to do this with examples

## Modifying factor order
If you want to explore the average number of hours spent watching TV per day across religions, we would run a query like:

```{r}
relig_summary <- gss18 |>
  filter(!is.na(relig)) |>
  group_by(relig) |>
  summarize(
    tvhours = mean(tvhours, na.rm = TRUE)
  )

ggplot(relig_summary, aes(x = tvhours, y = relig)) + 
  geom_point()
```

It is hard to read this plot because there’s no overall pattern. We can use `fct_reorder` to improve this.  It takes 3 arguments:

* .f, the factor whose levels you want to modify.
* .x, a numeric vector that you want to use to reorder the levels.


```{r}
ggplot(relig_summary, aes(x = tvhours, y = fct_reorder(relig, tvhours))) +
  geom_point()
```

Cleaner way of writing this would be:

```{r}
relig_summary |>
  mutate(
    relig = fct_reorder(relig, tvhours)
  ) |>
  ggplot(aes(x = tvhours, y = relig)) +
  geom_point()
```



## Exercise
Fix the below chart to make it more easily readable by showing the bars arranged from highest to lowest count, making the chart easier to interpret.


```{r bar-reorder, exercise = TRUE}
gss18 <- read_csv('data/gss18.csv')
gss18 |> count(income) |> ggplot(aes(y = income, x = n)) + geom_col()
```

```{r bar-reorder-solution}
gss18 |> count(income) |> ggplot(aes(y = fct_reorder(income, n), x = n)) + geom_col()
```


## Reordering on 2 variables

Another type of reordering is useful when you are coloring the lines on a plot. `fct_reorder2(.f, .x, .y)` reorders the factor `.f` by the `.y` values associated with the largest `.x` values. This makes the plot easier to read because the colors of the line at the far right of the plot will line up with the legend.

```{r}
by_age <- gss_cat |>
  filter(!is.na(age)) |> 
  count(age, marital) |>
  group_by(age) |>
  mutate(
    prop = n / sum(n)
  )

ggplot(by_age, aes(x = age, y = prop, color = marital)) +
  geom_line(linewidth = 1) + 
  scale_color_brewer(palette = "Set1")

ggplot(by_age, aes(x = age, y = prop, color = fct_reorder2(marital, age, prop))) +
  geom_line(linewidth = 1) +
  scale_color_brewer(palette = "Set1") + 
  labs(color = "marital") 
```

## `fct_infreq` and `fct_rev`

Use fct_infreq() to order levels in decreasing frequency.  To understand this concept we will use the `gss_cat` data that is provided by `tidyverse`


```{r}
gss_cat |>
  mutate(marital = fct_infreq(marital))  |>
  ggplot(aes(y = marital)) +
  geom_bar()
```

Combine it with fct_rev() if you want them in increasing frequency so that in the bar plot largest values are on the right, not the left.

```{r}
gss_cat |>
  mutate(marital = fct_rev(fct_infreq(marital))) |>
  ggplot(aes(y = marital)) +
  geom_bar()
```

## fct_recode

Change factor levels by hand using `fct_recode`

```{r}
x <- factor(c("apple", "bear", "banana", "dear"))
fct_recode(x, fruit = "apple", fruit = "banana")
```

##  `fct_collapse`

If you want to collapse a lot of levels, fct_collapse() is a useful variant of fct_recode(). For each new variable, you can provide a vector of old levels

```{r}
gss18 |>
  mutate(
    partyid = fct_collapse(partyid,
      "rep" = c("strong republican", "not str republican"),
      "ind" = c("ind,near rep", "independent", "ind,near dem"),
      "dem" = c("not str democrat", "strong democrat")
    )
  ) |>
  count(partyid)
```

## Lumping small groups

Sometimes you just want to lump together the small groups. `fct_lump_*()` family of functions can help. 

`fct_lump_lowfreq()`that progressively lumps the smallest groups categories into “Other”, always keeping “Other” as the smallest category.

```{r}
gss18 |>
  mutate(relig = fct_lump_lowfreq(relig)) |>
  count(relig)
```

However this reduced to only 4 categories.  Instead, we can use the fct_lump_n() to specify that we want exactly `n` groups:

```{r}
gss18 |>
  mutate(relig = fct_lump_n(relig, n = 10)) |>
  count(relig, sort = TRUE)
```

## More analysis
Let us analyse attitudes about religion

```{r}

# distribution of religion
gss18 |> ggplot(aes(x = relig, fill = relig)) +
  geom_bar()
```

How to improve this overlapping, cramped plot? Easy fix is to switch the axis

```{r}
gss18 |> ggplot(aes(y = relig, fill = relig)) +
  geom_bar()
```
Let us check the levels

```{r}
levels(gss18$relig)
```

Let us use `fct_infreq` to reorder the levels such in increasing order of frequency:

```{r}
gss18 |> ggplot(aes(y = fct_infreq(relig), fill = relig)) +
  geom_bar()
```

We may also want to simplify the plot by grouping some factors together. The fct_lump command will combine uncommon factor levels into a new level, "other":

```{r}
gss18 |> ggplot(aes(x = fct_lump(relig), fill = fct_lump(relig))) +
  geom_bar()
```

Be careful with this. Not everybody enjoys being lumped into the "Other" category. A good compromise in this case could be to use one bar for "Other", but still break everyone out in the legend:

```{r}
# lumped levels, but still showing the variation
gss18 |> ggplot(aes(x = fct_lump(relig), fill = relig)) +
  geom_bar()
```

## gss - generations
Here is the age distribution of 2014 surveyers

```{r}
gss18 |> ggplot() + geom_histogram(aes(x = age))
```

The oldest person was born in which year?

```{r oldest-person, exercise=TRUE}

```

```{r oldest-person-solution}
gss18 |> slice_max(age, n = 1) |> head(1) -> oldest_person
2018 - oldest_person$age
```

Let's define the following generations:

[Silent generation](https://en.wikipedia.org/wiki/Silent_Generation): 1929—1945
[Baby boomers](https://en.wikipedia.org/wiki/Baby_boomers): 1946—1964
[Gen X](https://en.wikipedia.org/wiki/Generation_X): 1964—1981
[Gen Y/Millennial](https://en.wikipedia.org/wiki/Millennials): 1982—1996
[Gen Z](https://en.wikipedia.org/wiki/Generation_Z): 1997—20teens

```{r}
gss18_gen <- mutate(gss18, birth_year = 2018 - age, 
                    generation = cut(birth_year, 
                                     breaks = c(1929, 1946, 1964, 1982, 1997, 2014), 
                                     labels=c('silent', 'boomer', 'x', 'millennial', 'z')))
```

```{r}
ggplot(gss18_gen) + geom_bar(aes(x = generation, fill = generation))
```

## Religious affiliation by generation
Here is the overall distribution of religious affiliation in the data:

```{r}
gss18_gen <- gss18_gen |> mutate(fct_lump_relig = fct_lump(relig))
gss18_gen |> select(generation, fct_lump_relig) |> table()
```

We probably want to normalize this table so that the data are expressed as percentages, with the total summing to one:

```{r}
gss18_gen |> select(generation, fct_lump_relig) |> table() |> prop.table()
```

```{r}
gss18_gen |> select(generation, fct_lump_relig) |> table() |> prop.table() |> sum()
```

Let us now transforms the row into a probability distribution where each row item represents a proportion of the total of sum 1

Note that by normalizing the rows or columns, we are now studying the conditional distribution of religion given generational status (or vice versa)

```{r}
# normalize the row

gss18_gen |> select(generation, fct_lump_relig) |> table() |> prop.table(1)
```

Next we will do conditional distribution of generation; given a religion

```{r}
gss18_gen |> select(generation, fct_lump_relig) |> table() |> prop.table(2)
```


## Hours of TV watched

```{r}
gss18_gen |> ggplot(aes(x = generation, y = tvhours)) + geom_boxplot()
```
This plot could maybe be improved by ordering the boxes so that the medians are in ascending order:

```{r}
gss18_gen |> ggplot(aes(x = fct_reorder(generation, tvhours), y = tvhours)) + geom_boxplot()
```

## Political orientation
```{r}
select(gss18_gen, generation, partyid) |> table() |> prop.table(1)
```





