---
subtitle: "DATASCI 306: Lecture 21"
title: "Shiny apps"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)     # For interactive exercises
library(tidyverse)
library(shiny)
```


## What is shiny?

* Shiny is an R package to create interactive web applications (apps)
* No need to know HTML, CSS, or JavaScript. 
* Can create interactive analytics dashboards  

[Get inspired by award winning dashboards!](https://medium.com/make-your-data-speak/myds-award-2023-oh-dashboards-my-dashboards-8c978fa2a468)

## Main building blocks of shiny

* UI - user interface that defines how your app looks
* Server function -  that defines how your app works

Shiny uses reactive programming to automatically update outputs when inputs change, so we’ll need the third important component of Shiny apps: **reactive expressions**. 

## Hello World app

There are several ways to create a Shiny app:

* The simplest is to create a new directory for your app, and put a single file called app.R in it. 

This app.R file will be used to tell Shiny both how your app should look, and how it should behave. Try it out by adding an app.R file that looks like this:

```{r, eval=FALSE, echo=TRUE}
ui <- fluidPage(
  "Hello, world!"
)

server <- function(input, output, session) {
}

shinyApp(ui, server)
```

This is a complete, if trivial, Shiny app! Looking closely at the code above, our app.R does four things:


- It defines the user interface, the HTML webpage that humans interact with. In this case, it’s a page containing the words “Hello, world!”.

- It specifies the behavior of our app by defining a server function. It’s currently empty, so our app doesn’t do anything, but we’ll be back to revisit this shortly.

- It executes `shinyApp(ui, server)` to construct and start a Shiny application from UI and server.

## Running

There are a few ways you can run this app:

- Click the Run App button in the document toolbar.

- Use a keyboard shortcut: Cmd/Ctrl + Shift + Enter



## Imperative vs declarative programming

* In `imperative programming`, you issue a specific command and it’s carried out immediately. 

* In `declarative programming`, you define your function, and rely on the framework to decide how and/or when to invoke your function

##  Reactive expressions

* They take inputs, which can be user-generated (e.g., slider values, text input) or other reactive expressions.
* They perform calculations or data manipulations based on these inputs.
* They produce an output, which can be used in other parts of the app, such as rendering plots, tables, or text.

## Reactivity

The key feature of reactive expressions is their reactivity.

* They are only executed when their input values change. This means that if an input remains unchanged, the expression is not re-evaluated, saving computational resources.

* When an input does change, the reactive expression is automatically re-executed, and its output is updated.

* This automatic updating mechanism ensures that the app always displays the most current information based on the user's interactions.

Some of the functions you can use to add reactivity:
`renderText()`,  `reactive()`,  etc.

## Work with a reactive programm

Let us run this program present in the `hello` directory under `shiny_example_apps` folder

```
ui <- fluidPage(
  textInput("name", "What's your name?"),
  textOutput("greeting")
)

server <- function(input, output, session) {
  output$greeting <- renderText({
    paste0("Hello ", input$name, "!")
  })
}
```
## Tutorial

Now let us run through some examples from this eBook:

https://mastering-shiny.org/basic-app.html

## Shiny cheatsheet

https://rstudio.github.io/cheatsheets/html/shiny.html

## Setting working directory

By default, the console tab opens the `/cloud/project/` folder.

You can check it by running the `getwd()` command

To change the working directory to some other directory use `setwd('directory_path')`

For example to change to `shiny-project-er-injuries` folder you would key in

`setwd('./shiny-project-er-injuries')` using relative path

or

`setwd('/cloud/project/lectures/shiny_example_apps/shiny-project-er-injuries')` using absolute path

## Example 1

Let us now understand the ncaa project. The code is part of this repository under 

shiny_example_apps/shiny-project-ncaa



## Example 2

shiny_example_apps/shiny-project-er-inuries


