---
subtitle: "DATASCI 306: Lecture 14"
title: "Strings, text data and date/time"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(nycflights13)
set.seed(2939394)
library(remotes)
# remotes::install_github("bradleyboehmke/harrypotter")
library(harrypotter)
```

## Today's lecture

* Understand the basics of working with strings in R.
* Convert data to strings.
* Extract data from strings.
* Using date/time

These notes correspond to Chapter 14 and 17 of your book.

## String basics

A string (or "character vector" in R) is character data enclosed in double `"` or single `'` quotes.

To include quotation marks inside, we need to **escape** them:
```{r}
mle <- "\"MLE\" stands for \"Maximum Likelihood Estimate\""
mle
cat(mle) # prints directly instead of displaying as escaped string
```

Which other way can you fix it?

```{r}
# another way of getting around this issue
'"MLE" stands for "Maximum Likelihood Estimate"'
```

## Other special characters

Since `\` is used to escape things, if you need a literal backslash use `\\`. 

Example:

New line: `\n` - use `\\n` to escape the backslash if you want `\n` to be part of the string

```{r}
cat('first line \n second line')

cat('first line \\n not second line')
```



## ASCII and Unicode

Early computers could only read and write the ASCII character set, essentially just roman letters, numbers and some punctuation.

Nowadays, computers need to be able to understand alphabets from all over the world. For this we have Unicode.

You can print characters if you know their unicode using `\u`. For example, the copyright character has unicode 00A9. Wikipedia has a [complete list](https://en.wikipedia.org/wiki/List_of_Unicode_characters).

```{r}
print("\u00A9")
```

## String functions in R
Through the `stringr` package, `tidyverse` has a lot of useful functions for combining strings.

## Combining strings

If we need to make strings out of other strings, we have two (more or less identical) functions:

* Old school: `paste` (joins with " ") and `paste0` (joins with "");
* New school: `str_c` (joins with ""); used in our book

```{r}
paste('a','b','c')

paste("Section", c("100", "200", "100"), 
      "meets on", 
      c("Tuesday", "Monday", "Wednesday"))
```
Same function using `str_c`


```{r}
str_c('a','b','c')

str_c("Section", c("100", "200", "100"), 
      "meets on", 
      c("Tuesday", "Monday", "Wednesday"))
```
To separate the strings, use `sep` keyword argument

```{r}
str_c('a','b','c', sep = ", ")

str_c("Section", c("100", "200", "100"), 
      "meets on", 
      c("Tuesday", "Monday", "Wednesday"), sep = " ")

```

## Exercise

Join the following using a new line characters. Demonstrate using `str_c`.

```{r sonnet18, exercise = TRUE}
line1 <- "Nor shall death brag thou wander'st in his shade,"
line2 <- "When in eternal lines to time thou grow'st:"
line3 <- "So long as men can breathe or eyes can see,"
line4 <- "So long lives this, and this gives life to thee."
```

```{r sonnet18-solution}
line1 <- "Nor shall death brag thou wander'st in his shade,"
line2 <- "When in eternal lines to time thou grow'st:"
line3 <- "So long as men can breathe or eyes can see,"
line4 <- "So long lives this, and this gives life to thee."
cat(str_c(line1, line2, line3, line4, sep = '\n'))
```


## Properties of strings

```{r, echo = FALSE}
line1 <- "Nor shall death brag thou wander'st in his shade,"
line2 <- "When in eternal lines to time thou grow'st:"
line3 <- "So long as men can breathe or eyes can see,"
line4 <- "So long lives this, and this gives life to thee."
```

The `length` function in R will tell you how many items are in a vector:

```{r}
sonnet18_end <- c(line1, line2, line3, line4)
length(sonnet18_end)
```

To find the length of the strings in terms of characters:
```{r}
str_length(sonnet18_end) # base::nchar also works
```

## Flattening strings

`str_flatten` will take a vector of strings and flatten them into one string.

```{r}
# list all the air carriers flying out of each airport
distinct(flights, origin, carrier)
distinct(flights, origin, carrier) |>
  group_by(origin) |>
    summarize(carrier = str_flatten(carrier, ", "))
```



## Subsetting

The `str_sub` function can retrieve or change **substrings**.

```{r}
line1
str_sub(line1, 11, 20)
line2
str_sub(line2, 36)
str_sub(line2, -8)
line4
line4b <- line4
str_sub(line4b, 1, 10) <- "REMOVED"
line4b
```


## Splitting

Here's the full sonnet as a single string:
```{r}
sonnet18 <- "Shall I compare thee to a summer's day?\nThou art more lovely and more temperate:\nRough winds do shake the darling buds of May,\nAnd summer's lease hath all too short a date;\nSometime too hot the eye of heaven shines,\nAnd often is his gold complexion dimm'd;\nAnd every fair from fair sometime declines,\nBy chance or nature's changing course untrimm'd;\nBut thy eternal summer shall not fade,\nNor lose possession of that fair thou ow'st;\nNor shall death brag thou wander'st in his shade,\nWhen in eternal lines to time thou grow'st:\n\tSo long as men can breathe or eyes can see,\n\tSo long lives this, and this gives life to thee."
```

Let's split that up into individual lines as strings in a vector:

```{r}
sonnet18_lines <- str_split(sonnet18, "\n") # notice use of new line
sonnet18_lines
```


## `unlist` function

`unlist` takes a list (or a vector of lists) as input and converts it into a vector. It essentially flattens the list structure, removing any nested lists and combining all elements into a single vector

Let us take a simple character vector to understand `unlist` function 

```{r}

x <- list('abc','xyz')
x[1]
x[[1]]

z <- unlist(x)
z[1]
z[[1]]
```


## Exercise

**Tokenization** is the process many computer languages use to turn strings into discrete words. Use `str_split` to turn each line of `sonnet18` into individual words. Use `unlist` on the result and answer the question: how many words are there?

```{r tokenize1, exercise = TRUE}
sonnet18 <- "Shall I compare thee to a summer's day?\nThou art more lovely and more temperate:\nRough winds do shake the darling buds of May,\nAnd summer's lease hath all too short a date;\nSometime too hot the eye of heaven shines,\nAnd often is his gold complexion dimm'd;\nAnd every fair from fair sometime declines,\nBy chance or nature's changing course untrimm'd;\nBut thy eternal summer shall not fade,\nNor lose possession of that fair thou ow'st;\nNor shall death brag thou wander'st in his shade,\nWhen in eternal lines to time thou grow'st:\n\tSo long as men can breathe or eyes can see,\n\tSo long lives this, and this gives life to thee."
```

```{r tokenize1-solution}
str_split(sonnet18, " ") |> unlist() |> length()
```

## separate_ functions
`tidyverse` has some built-in functions designed to break apart and combine string data:

* `separate_longer_delim`, `separate_longer_position`: combine multiple columns of string data into one.
* `separate_wider_delim`, `separate_wider_position`: extract string data into multiple columns. These functions work analogously to pivot_{wider,longer}.

`separate_wider_delim` converts a delimited string into multiple columns of data. separate_wider_position does the same, but uses a fixed position


```{r}

df <- tibble(x = c("a10.1.2022", "b10.2.2011", "e15.1.2015"))
df |>  separate_wider_delim(
    x,
    delim = ".",
    names = c("code", "edition", "year")
  )
```

## Harry Potter

Let's practice some string analysis by looking at the text of Harry Potter:

<img src = 'images/harry-potter.jpeg' />

`philosophers_stone` is a list containing 17 entries: one for each chapter of "Harry Potter and the Sorcerer's Stone" (titled differently in the U.K.). This object is created in the initialization block of this page

Instead of reading Harry Potter, we will analyze it like a data set.

Each entry of `philosophers_stone` is a string. There is one string per chapter, so they are pretty long. 


```{r}
ch1 <- philosophers_stone[[1]]
length(philosophers_stone)
ch1
```

## Exercise
What is the longest (in number of characters) chapter in Philosopher's Stone?

```{r longest-chapter, exercise = TRUE}
philosophers_stone
```

```{r longest-chapter-solution}
str_length(philosophers_stone) |> max()
```

## `str_count`

Use this function to count the number of times a `pattern` is found within the string

```{r}
# How many times Harry occurs in chapter 1?
str_count(ch1, 'Harry')
```

## `str_locate`

returns the start and end position of the first match; `str_locate_all()` returns the start and end position of each match.

```{r}
str_locate(ch1, 'Harry')
```

Let is verify

```{r}
str_sub(ch1, 5243, 5247)
```

## Exercise
What are the last 100 words of ch1?

```{r last1-100, exercise=TRUE}
ch1 <- philosophers_stone[[1]]
```

```{r last1-100-solution}
ch1 |> str_sub(length(ch1)-101) 
# another way
ch1 |> str_sub(-100)
```

## Exercise
What is the last sentence of ch1? 
Hint: `str_split` might be useful. Since a period `.` is a special character, you have to use the expression `\\.` to use it in functions. More on this in the next class

```{r last-sentence, exercise = TRUE}

```

```{r last-sentence-solution}
ch1 <- philosophers_stone[[1]]
ch1 |> str_split("\\.") -> sentences 
sentences[[length(sentences)]][length(unlist(sentences))]
```

## Exercise
Which chapter starts with the word `THE POTIONS MASTER`?

```{r starts-with, exercise = TRUE}

```

```{r starts-with-solution}
str_starts(philosophers_stone, 'THE POTIONS MASTER')
```


## Converting the list to tibble

We may find it useful sometimes to convert the list to a tibble and then apply `tidyverse` functions. Here is the previous solution using a tibble


```{r}
df <- tibble(philosophers_stone)
df |> mutate(starts_with = str_starts(philosophers_stone, 'THE POTIONS MASTER')) |> select(starts_with) |> print()
```

## Dates 
Most of us have a pretty firm grasp on dates. But they can be more complicated than you might think. Consider these three questions from the book:

* Does every year have 365 days?
* Does every day have 24 hours?
* Does every minute have 60 seconds?

The answer to all three is "no".

## Date - Types

There are three different date classes in R:

* A date, printed in tibble as `<date>`, represents a full day on the calendar.
* A time within a day, printed as `<time>`, represents a specific time within an (unspecified) day.
* A date-time is a date plus a time (tibble: `<dttm>`). A date-time uniquely identifies an instant in time (up to a given precision, usually 1 second.)

There are lots of different formats, timezones, leapyears to deal with. The `lubridate` package helps:

```{r}
library(lubridate)
```


## Dates and date-times
Some useful functions

```{r}
today()
now()
```

We’ve already seen examples of date-times in the flights tibble:

```{r}
select(flights, time_hour) |> print()
```

## The `mdy`, `ymd`, and `dmy` functions

Converting strings in different formats, to dates and times, is easy with these convenience functions:

```{r}
# Only date
mdy("10/01/2022") # Oct. 1
ymd("2022-10-01")
dmy("01.10.2022")
dmy("31-Jan-2017")
dmy("31 1 2017")

# Date and time
ymd_hms("2017-01-31 20:11:59")
mdy_hms("10/01/2022 12:01:37")
mdy_hm("01/31/2017 08:01")

```
`UTC` stands for Coordinated Universal Time. 
It is a standard used to establish time zones worldwide.
EST is in the UTC-5 time zone


`_hms` version will parse `HOURS:MINUTES:SECONDS` while `_hm` will just be `HOURS:MINTUES`. AM/PM can be included, otherwise 24 clock assumed. 

## Setting Timezone

We can select timezone using the nearest major city:
```{r}
mdy_hms("10/01/2022 12:01:37", tz = "America/Detroit")
mdy_hms("12/01/2022 12:01:37", tz = "America/Detroit")
```

## Exercise

Parse the following into dates or date times:
```{r parsed, exercise = TRUE}
dt1 <- "31/10/89"
dt2 <- c("2022-07-01 3:10am",
         "1852-01-15 7:00pm")
```


```{r parsed-solution}
dt1 <- "31/10/89"
dt2 <- c("2022-07-01 3:10am",
         "1852-01-15 7:00pm")
dmy(dt1)
ymd_hm(dt2)
```

## using `as_datetime`

Sometimes you can get away with calling as_datetime()

```{r}
as_datetime("2017-01-02 20:11:59")
```

But not always
```{r}
as_datetime("01-02-2017 20:11:59")
```

## `parse_datetime`
Another convenience function to use when as_datetime fails

```{r}
parse_datetime("22/10/1, 10:01:00", 
               format = "%y/%m/%d, %H:%M:%S")
```

>* Year: four digit `"%Y"`, two digit `"%y"`
>* Month: number `"%m"`, abbreviation `%b`, full name `%B`
>* Day: number in month `"%d"`
>* Hours:minutes:seconds: `"%H:%M:%S"`
>* See `parse_datetime` documentation for more

## using `make_date` and `make_datetime`

These functions can make dates from individual (numerical) components:

```{r}
make_datetime(2024, 3, 5, 10, 00, 59, tz = "EST")
make_date(2024, 01, 10)
```

## The epoch

UNIX systems sometimes represent time as "the number of seconds which have elapsed since 1-1-1970." This date is known as "the epoch". So you may occasionally come across date-times that look like:

```{r}
now() |> as.integer()
```
To convert these to date format you can use as_datetime:

```{r}
as_datetime(3600 * 24)  # 24 hours after the epoch
```





## Extracting components

```{r}
right_now <- now()
day(right_now) # same as mday()
month(right_now)
year(right_now)
yday(right_now) # day within year ("Julian date")
wday(right_now) # day within week, starting with Sunday
hour(right_now)
minute(right_now)
second(right_now)
```

## Time Math 

```{r}
ymd("2023/01/04") - mdy("12/16/2022")
```

The `seconds`, `minutes`, `hours`, `days`, `weeks` can be used to get offsets:
```{r}
right_now <- now()
right_now
right_now + minutes(10)
right_now + weeks(7)
```

## Time spans

After doing math, you might want to convert the units.

```{r}
# Days to Last day of class!
semester_end <- dmy_hms("09122024 23:59:59") - now()
# difftime object is returned

sed <- as.duration(semester_end)

as.integer(sed, "days")
as.numeric(sed, "minutes")

```
You may use update function to update a specific part of the date

```{r}
(d = today() - update(today(), day = 26))
```

## Durations
There are equivalent `d*` functions (`dyears`, etc) for constructing durations directly.

Any time part can be converted to seconds using duration class

```{r}
as.duration(1234)
dseconds(10)
dweeks(3)
```


## Arithmetic with duration

Arithmetic with durations works as you would expect:

```{r}

2 * dyears(1)
dyears(1) + dweeks(12) + dhours(15)
```

You have to be careful when adding durations and dates. What is 1 day after March 8th, 2025 at 1pm?

```{r}
# But what happened to this addition?
march8 = mdy_hm("March. 8, 2025 13:00PM", tz="America/Detroit")
march8 + ddays(1)

# this seems fine
march9 = mdy_hm("March. 9, 2025 13:00PM", tz="America/Detroit")
march9 + ddays(1)
```

You can use methods of periods from lubridate instead

```{r}
march8 = mdy_hm("March. 8, 2025 13:00PM", tz="America/Detroit")
march8 + days(1)

```
## Exercise

Parse these dates.

* What's the range of the dates? 
* What is the mean date?
* What is the median date? 
* Suppose you want reminders of the anniversary of these dates in the coming year one month in advance. Find the dates of the reminders.

```{r diffex, exercise = TRUE}
ds <- c("1981-09-25", "1982-03-06", "2010-12-14", "2012-10-18", "2019-04-03")
```

```{r diffex-solution}
ds <- c("1981-09-25", "1982-03-06", "2010-12-14", "2012-10-18", "2019-04-03")

# parse as year-month-day format
ds_parsed <- ymd(ds)

# range
range(ds_parsed)
# or
min(ds_parsed)
max(ds_parsed)

# mean and median
mean(ds_parsed)
median(ds_parsed)

# setting reminders
## method 1
year_diff <- year(today()) - year(ds_parsed)
ds_this_year <- ds_parsed + years(year_diff) - months(1)
if_else(ds_this_year < today(), ds_this_year + years(1), ds_this_year) 

## method 2
ds_parsed_2 <- ds_parsed - months(1)
year(ds_parsed_2) <- year(today())
ds_parsed_2 + if_else(ds_parsed_2 < today(), years(1),  years(0))
```



## Revisiting of some common operations on Date columns

```{r}
aatemp <- read_csv("data/ann_arbor_weather.csv.gz")
aatemp |> select(DATE) |> head(3)


## Finding maximum TMAX per year-month 
aatemp |> group_by(year(DATE), month(DATE)) |> 
  summarize(m = max(TMAX, na.rm = TRUE)) |>
  arrange(desc(m)) |> head(5)

## limiting to a specific date range
aatemp |> filter(DATE >= "2018-01-01", DATE <= "2018-03-31") |>
  nrow() # 31 + 28 + 31
```

## Exercise

Using the `weather` table in the `nycfights13` package, compute the average temperature at each airport (`origin`) between 9am and 15pm on September 1, 2013 (see the `time_hour` column).

```{r nyc0901temp, exercise = TRUE}

```


```{r nyc0901temp-solution}

filter(weather, time_hour >= "2013-09-01 09:00:00", time_hour <= "2013-09-01 17:00:00") |>
  group_by(origin) |> summarize(mean(temp))
```

## Exercise

The `nyc flights` table has scheduled as well as actual arrival and departure times. Let's create a date-time variable from the actual departure time.

First we need to take note of the fact that dep_time and arr_time are in a weird format:

```{r}
flights |> select(dep_time, sched_dep_time, arr_time, sched_arr_time, year, month, day) |> head()
```

Now convert the values to dttm

```{r nycdate, exercise = TRUE}

```

```{r nycdate-solution}
flights_dt <- flights |>
  mutate(across(c(dep_time, sched_dep_time, arr_time, sched_arr_time), ~ make_datetime(year, month, day, . %/% 100, . %% 100)))
```
## Analysis

Now we can use built in R command to query and plot these data based on actual departure time

```{r}

flights_dt <- flights |>
  mutate(across(c(dep_time, sched_dep_time, arr_time, sched_arr_time), ~ make_datetime(year, month, day, . %/% 100, . %% 100)))

flights_dt |>
  filter(dep_time < ymd(20130130), dep_time >= ymd(20130120)) |>
  ggplot(aes(x = dep_time)) +
  geom_density(bw = 3*3600)
```

**Histogram**

Let us also plot a histogram 

```{r}
flights_dt |>
  filter(dep_time < ymd(20130130), dep_time >= ymd(20130120)) |>
  ggplot(aes(x = dep_time)) +
  geom_histogram()
```

Notice that this command and resulting graph looks much nicer than if we did not use the date classes:

* The filter on dep_time looks more natural than filter(day < 2, month == 1, year == 2013).
* We can easily express the bin-width of 600 seconds in geom_freqpoly(binwidth = 600). When you use date-times in a numeric context (like in a histogram), 1 means 1 second; for dates, 1 means 1 day.
* The plot x axis has nice readable labels.

## More analysis

By extracting the minute component of scheduled departure time, we uncover a surprising pattern:

```{r}
flights_dt |>
  mutate(minute = minute(dep_time)) |>
  group_by(minute) |>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE)) |>
  ggplot(aes(minute, avg_delay)) +
    geom_line() + labs(x="minute(dep_time)")
```

But when grouped by scheduled departure, time the delays seem to be random:

```{r}
sched_dep <- flights_dt |>
  mutate(minute = minute(sched_dep_time)) |>
  group_by(minute) |>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n = n())

ggplot(sched_dep, aes(minute, avg_delay)) +
  geom_line()
     
```

