---
title: "DATASCI 306: Lecture 12"
subtitle: "SQL"
output:
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(DBI)
library(dbplyr)
library(tidyverse)
library(nycflights13)
library(lubridate) # install.packages("lubridate") if you don't have this
```

## SQL - Structured Query Language

A huge amount of data lives in relational databases so it is important to understand how to connect to a relational database and work with it.

Today's lecture corresponds to Chapter 21 of your book

## Database Connection

To connect to the database from R, you’ll use a pair of packages:

* DBI (database interface) - this provides a set of generic functions that connect to the database, upload data, run SQL queries, etc.
* You’ll also use a package tailored for the DBMS you’re connecting to. This package translates the generic DBI commands into the specifics needed for a given DBMS. There’s usually one package for each DMBS, e.g. RPostgres for Postgres and RMariaDB for MySQL. In this example we use SQLite and the package to use is RSQLite

## Let's create a database

```{r}

# Create an ephemeral in-memory RSQLite database
con <- dbConnect(RSQLite::SQLite(), ":memory:")

dbListTables(con)
```

At the outset you see that there are no tables when you run the above command.

## dbplyr

dbplyr is a dplyr backend, that allows you to keep writing dplyr code and dbplyr translates it to SQL. Now we are going to create a table called 'mpg' in our SQLite database using the mpg tibble and dbplyr is generating the necessary SQL queries to get this done behind the scenes.

```{r}
dbWriteTable(con, "mpg", mpg, overwrite = T)
dbListTables(con)
```

To remove

```{r}
# dbRemoveTable(con, 'mpg') # this will delete the table
```

Now let us look into the column names of this table

```{r}
dbListFields(con, "mpg")
```

Time to read the entire table!

```{r}
dbReadTable(con, "mpg")
```

## Selecting data from a table

The SQL syntax for selecting column(s) from a table is


    sql
    SELECT <col1>, <col2>, ..., <coln> FROM <table>


Note the similarity to the corresponding tidyverse command:

    r
    select(<table>, <col1>, <col2>, ..., <coln>)


Let us select some records

```{r}
res <- dbSendQuery(con, "SELECT * FROM mpg")
res
```

```{r}
dbFetch(res) |> head()
```

The special keyword * means "select everything" and is equivalent to dplyr's everything():

If you have a really big table, SQL allows you to LIMIT the number of rows it returns.

```{r}
# using limit
res <- dbSendQuery(con, "SELECT * FROM mpg LIMIT 2")
dbFetch(res)
```

## Filtering

The SQL syntax for filtering rows in a table uses the WHERE clause:


    sql
    SELECT * FROM <table> WHERE dest = "IAH"


This is the same as:


    r
    filter(<table>, dest == "IAH")


Note that SQL uses a single = to check equality!

```{r}
res <- dbSendQuery(con, "SELECT * FROM mpg WHERE cyl = 4")
dbFetch(res)
```

## More tables

Now let us create more tables using our tidyverse datasets

```{r}
dbWriteTable(con, "flights", flights, overwrite = T)
dbWriteTable(con, "airports", airports, overwrite = T)
dbWriteTable(con, "airlines", airlines, overwrite=T)
dbWriteTable(con, "planes", planes, overwrite = T)
dbWriteTable(con, "weather", weather, overwrite = T)
```

We will define a convenience function to simplify our query

```{r}
q <- function(...) dbGetQuery(con, ...)
print(q)
```

## Missing data
In SQL, missing data is coded as NULL. This is a special value which is 
analogous to NA in R.

```{r}
# select non-null dep time

q('SELECT * FROM flights WHERE dep_delay IS NOT NULL') |> as_tibble() |> print()

# filter(flights, !is.na(dep_delay))
```

## Summarizing

The SQL syntax for summarizing is using the `GROUP BY` clause:

    {sql}
    SELECT AVG(<col>) AS avg_col FROM <table> GROUP BY(<group cols>)
    
This is the same as:

    {r}
    <table> |> group_by(<group cols>) |> summarize(avg_col = mean(<col>))
    
```{r}
# select the average distance and count for each day

q("SELECT year, month, day, AVG(distance) AS avg_dist,
    COUNT() AS n FROM flights GROUP BY year, month, day")

# equivalent to
# flights |> group_by(year, month, day) |> summarize(avg_dist = mean(distance), n = n())
```

## Joins

The SQL syntax for joins:

    {sql}
    SELECT * FROM <table> LEFT JOIN <other_table> ON <left_key_col> =            <right_key_col>

This is the same as:

    {r}
    left_join(<table>, <other_table>, by = c("<left_key_col>" =         "<right_key_col>"))


```{r}
q("SELECT flights.year, planes.tailnum AS flightyr, * FROM flights
    LEFT JOIN planes ON flights.tailnum = planes.tailnum")
```


## Ranking

Ranking operates slightly differently in SQL than dplyr. The overall syntax is:

    SELECT RANK() OVER (
        PARTITION BY <columns>
        ORDER BY <expression1> [ASC|DESC], ...
    ) RankedCol FROM <table>
    
```{r}
# rank departure delay by month, day

q('SELECT year, month, day, dep_delay, RANK() OVER (
    PARTITION BY month, day
    ORDER BY dep_delay DESC
    ) r FROM flights')
```

The above query is equivalent to

```{r}
flights |> group_by(month, day) |> mutate(r = min_rank(-dep_delay)) |>
    select(month, day, dep_delay, r) |> arrange(month, day, r)
```

## Subqueries
An aspect of SQL I find confusing is that you cannot filter (WHERE ...) based on a ranking condition:

```{r eg, exercise = TRUE}
q('SELECT year, month, day, dep_delay, RANK() OVER (
    PARTITION BY month, day
    ORDER BY dep_delay DESC
    ) r FROM flights where r  = 1')
```

You need to use subquery to get the result as shown below

```{r}

q('SELECT year, month, day, dep_delay, r FROM
    (
        SELECT year, month, day, dep_delay, RANK() OVER (
        PARTITION BY month, day
        ORDER BY dep_delay DESC
    ) r FROM flights)
    WHERE r <= 2
')
```

## Exercise

Find the max highway miles across each manufacturer


```{r}

res <- dbSendQuery(con, "select manufacturer, max(hwy) from mpg group by manufacturer")
dbFetch(res)
dbClearResult(res)
```

## Exercise
How many 4 cylinder models are present in this database?

```{r ex-4, exercise = TRUE}

```


```{r ex-4-solution}
res <- dbSendQuery(con, "select count(*) from mpg where cyl = 4")
dbFetch(res)
dbClearResult(res)
```

## Exercise

Extend the previous query to retrieve the minimum hwy value across manufacturer and model

```{r ex-5, exercise = TRUE}

```

```{r ex-5-solution}
res <- dbSendQuery(con, "SELECT manufacturer, model, min(hwy) FROM mpg GROUP BY manufacturer, model")
dbFetch(res)
dbClearResult(res)
```

## Using `having` clause

The `WHERE` keyword works with select. However, when you want to sub select based on grouped data, you have to use `HAVING` instead

Here is an example; Get all destinations that have total flights to that destination greater than 10000

```{r}

q('select dest, count(*) as total from flights group by dest having total > 10000' )
```

## Exercise
Get all the manufacturers who have 4 or more 'compact' car class

```{r ex-6, exercise = TRUE}

```

```{r ex-6-solution}
res <- dbSendQuery(con, "SELECT manufacturer, class, count(*) FROM mpg WHERE class = 'compact' GROUP BY manufacturer HAVING count(*) >= 4")
dbFetch(res)
dbClearResult(res)
```

## Exercise
Get distinct car manufacturers in this dataset

```{r ex-7, exercise = TRUE}

```

```{r ex-7-solution}
res <- dbSendQuery(con, "SELECT DISTINCT manufacturer FROM mpg")
dbFetch(res)
dbClearResult(res)
```

Once done, it is a good idea to disconnect from the database

```{r}
dbClearResult(res)
dbDisconnect(con)
```
## Other resources
* [w3schools](https://www.w3schools.com/sql/default.asp)
* [SQL murder myster](https://mystery.knightlab.com/)