---
title: "DATASCI 306: Lecture 3"
subtitle: "More data exploration with ggplot"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
```

## Review

* Learned more about R's functions (required and optional arguments); `f(x, y = 10)`
* Piping and assignment; `|>, =`
* Special values: `NA`, `NULL`, and `NaN`
* Vectors: ordered collections of same type, many computations are *vectorized*; See example below
* Lists: ordered collections of mixed types; `c(1:10); list(1, c(1:10))`
* Tables: lists of vectors of the same length (column oriented); `tibble, data.frame`
* Other items: indexing with `[]` and `$`, `dim` and `colnames` for tables

**Example On Vectorized operations**

Multiply all elements by 10

```{r vectorized, exercise = T}
v = c(1:10)
print(v)

```

## Plotting: Visual representation of data

**Graphing** or **plotting** is the representation of data in a visual form, typically on a 2D plane.

Humans have a wonderful ability to process visual stimuli quickly.

Great for displaying large data sets that might be difficult to describe.

**But Plotting ain't easy**

* What exactly can we put in a plot?
* How to connect the plot to data?
* What are our goals for the plot?
* What additional information is necessary to understand the plot?

## The Semiology of Graphics

In 1967, **Jacques Bertin** published the *Semiology of Graphics* to describe common elements of plots and what they could achieve.

Bertin described two ways of thinking about plots:

1. Visual ("retinal") variables: connections between objects in the plot and underlying data
2. Relationship what types of relationships can the visual variables express


[Visual Variables](https://en.wikipedia.org/wiki/Visual_variable)



## The Grammar of Graphics and `ggplot`

A follow up to Bertin's work was *The Grammar of Graphics* by Leland Wilkinson.
This book described a programming language for graphics based on ideas in
Bertin's system.

The GoG was implemented for R in `ggplot` (and later replaced by `ggplot2`). 

## The main components of a ggplot graph

>* The graph object itself (creating using `ggplot(data)`)
>* A set of [aesthetic mappings](https://ggplot2.tidyverse.org/articles/ggplot2.html?q=scales#mapping) (connecting data to visual variables)
>* [Layers](https://ggplot2.tidyverse.org/articles/ggplot2.html?q=scales#layers): collections of geometric elements (`geom_*()`) and [statistical transformations](https://ggplot2.tidyverse.org/reference/layer_stats.html) (`stat_*()`). Stat performs computation on the data before it is displayed
>* [Scales](https://ggplot2.tidyverse.org/articles/ggplot2.html?q=scales#scales): information on the range, breaks for labels and legends
>* [Coordinate systems](https://ggplot2.tidyverse.org/articles/ggplot2.html?q=scales#coordinates): how the data are arranged spatially
>* [Facet](https://ggplot2.tidyverse.org/articles/ggplot2.html?q=scales#facets): breaking a single plot into many, similar plots
>* [Theme](https://ggplot2.tidyverse.org/articles/ggplot2.html?q=scales#theme): all the other color and printing aspects of the plot

## Creating a ggplot

Start use the `ggplot` function to start our plot
```{r}
efficiency <- ggplot(data = mpg)
efficiency # for now, blank
```



**What can we plot?** Let us take a glimpse of the dataset
```{r}
glimpse(mpg)
```

"Aesthetic" mappings connect columns to visual variables
```{r}
efficiency <- ggplot(data = mpg,
                     aes(x = displ, 
                         y = hwy, 
                         color = cyl))

efficiency
```
displ: engine displacement is the measure of the total volume of all the cylinders in an internal combustion engine.
hwy: highway miles/gallon
cyl: number of cylinders


But the chart is still blank? Well we need to add the geometric layer to define how to display the visual elements

## Geometries: objects on the plot

We will use a **geometry function** (have the form `geom_TYPE()`).

```{r}
efficiency + geom_point()
```


**Alternative forms**

```{r eval = F}
ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy, color = cyl))

mpg |> ggplot() + geom_point(aes(x = displ, y = hwy, color = cyl))

mpg |> ggplot(mapping = aes(x = displ, y = hwy, color = cyl)) + geom_point()

ggplot(mpg, aes(x = displ, y = hwy, color = cyl)) + geom_point()

ggplot(mpg, aes(x = displ)) + geom_point(aes(y = hwy, color = cyl))
```

Note: You can use positional arguments or use `data`, `mapping` argument names to call these functions

Mapping values from `ggplot` are inherited by layers


## Question
Is the cylinder really a continuous variable?
How else can I make this better?

```{r ex1, exercise = TRUE}
mpg |> ggplot() + geom_point(aes(x = displ, y = hwy, color = cyl))

```


## Exercise: Great Lakes water levels

Here's a data set from the US Army Corps of Engineers showing the height of the Great Lakes (with Michigan and Huron being one body) above sea level, aggregate by year.

```{r}
glwl <- read_csv("data/GLHYD_data_english_cleaned.csv.gz") |>
  group_by(body, year) |> summarize(ht = median(height_ft)) |> ungroup()
glwl
```

Use this data set to make a scatter plot of the height of each lake in each year, colored by which `body` the reading is from.

```{r glwl-plot-setup}

glwl <- read_csv("data/GLHYD_data_english_cleaned.csv.gz") |>
  group_by(body, year) |> summarize(ht = median(height_ft)) |> ungroup()
```

```{r glwl-plot, exercise = TRUE}

```

Improve your plot by changing it to a `geom_line()`

## What aesthetic mappings can I use?


**Trying out some other mappings**

```{r}
ggplot(data = mpg,
       aes(x = displ, y = hwy, size = cyl, color = class)) + 
  geom_point()
```

Visit the docs to learn more: [geom_point](https://ggplot2.tidyverse.org/reference/geom_point.html)

## Using expressions

We can also use expressions involving columns.

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, shape = year > 2000)) + 
  geom_point()
```

## Overriding parameters for all points

We can pass in constants that apply to all points (size and transparency):

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, shape = year > 2000)) + 
  geom_point(size = 5)
```

## Jitter: useful noise

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) + 
  geom_point(position = "jitter")
```



## General observations

* Generally limit plots to having 3 or 4 distinct visual variables
* Almost everything can be tweaked in ggplot, finding it is the tricky part
* The [R Graph Gallery](https://r-graph-gallery.com/) is a great source of inspiration and instruction






## Other geometries

Points are great for graphing two **quantitative** or **continuous** variables.
For **discrete** or **qualitative** we need other charts.

```{r}
ggplot(data = mpg,
       aes(x = displ, y = class)) + 
  geom_boxplot() 
```



## List of geometries

[More ggplot documentation](https://ggplot2.tidyverse.org/reference/index.html#geoms).

## Statistical Summaries

* In addition to the raw data (or our calculations), our plots involved **data summaries** 
* `ggplot2` calls these **summary statistics** or `stat_*` functions
* We already saw a summaries in the boxplot: quantiles, twice IQR bars
* We can access summaries that geometries compute and add additional summaries.

## Boxplot as statistic

```{r}
ggplot(data = mpg,
       aes(x = displ, y = class)) + 
  stat_boxplot() 
```

Compare with

```{r}
ggplot(data = mpg,
       aes(x = displ, y = class)) + 
  geom_boxplot() 
```

## Adding computed summaries

The `stat_summary` function allows you to use any function to summarize 

```{r}
ggplot(data = mpg,
       aes(x = displ, y = class)) + 
  geom_boxplot() +
  stat_summary(fun = mean, size = 3, color = "red", geom = "point") 
```

## Trend lines

When using scatter plots, one of the most common summaries is a **trend line**.
```{r}
ggplot(data = mpg,
       aes(x = displ, y = hwy)) + 
  geom_point(position = "jitter", alpha = 0.25) +
  stat_smooth() # geom_smooth also works
```

## More layering

```{r}
ggplot(data = mpg, aes(x = displ)) +
  geom_point(aes(y = hwy), color = "orange") +
  geom_point(aes(y = cty), color = "blue") +
  stat_smooth(aes(y = hwy), lty = 1, color = "black") +
  stat_smooth(aes(y = cty), lty = 2, color = "red")
```

We'll see a better way to make this table when we talk about tall vs. wide format data.

## Overriding defaults of `stat` functions

Each `geom_*` has a default statistic function. We can override this.

```{r}
ggplot(data = mpg, aes(x = class)) +
  geom_bar() # default stat is count
```

Override the default

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  geom_bar(stat = "summary", fun = "mean") +
  labs(title = 'Mean highway miles per class')
```

## Replacing tables

We often use **tables** in documents to give numerical summaries. But why not
replace those with a nice graphic?

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  stat_summary(
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

## Exporting Plots

* Automatically included in knitted RMarkdown documents
* Export from the Plots panel in RStudio
* `ggsave` function to save to file (variety of formats)

