---
title: "DATASCI 306: Lecture 6"
subtitle: "More Grouping and Summaries; Functions"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(nycflights13)
library(lubridate) # install.packages("lubridate") if you don't have this
aatemp <- read_csv("data/ann_arbor_weather.csv.gz")
```

## Review: `mutate`

* `mutate`: Takes a table and returns a new table with columns added and/or subtracted
* `mutate(d, new_col = f(x, y), new_col2 = new_col1 + 1, old_col = NULL)`
* Can use functions that summarize, e.g., `mutate(d, x_centered = x - mean(x))`

## Today we will learn `groups and summaries`

* We often want to **partition** our data into smaller groups
* Things like finding **conditional means** or **conditional medians**
* `group_by` takes one or more grouping factors and creates a grouped table
* `summarize` computes group level summaries
* We'll see that `mutate` operates within groups, which can be quite helpful

## `nrow` function

Before we get started, let's look at a function to quickly count the rows in a table.

```{r}
mpg |> nrow()
```

Compare this with `dim`

```{r}
dim(mpg)
```
## Grouping

Often we want to break data out across categories and compute summaries within each.

```{r}
flights |> group_by(carrier) |>
  summarize(avg_dep_delay = mean(dep_delay, na.rm = T), total_flights = n())
```
Verify the slice

```{r ex-1, exercise = T}
flights |> 
  filter(carrier == '9E') |> 
  summarize(mean_delay = mean(dep_delay, na.rm = T), n = n())
```

## Using count
Many a times just using `count` function is enough instead of using `group_by` and summarize with `n()` function

```{r}
flights |> group_by(carrier) |> summarize(n = n())
```

The same result can be achieved by

```{r}
flights |> count(carrier)
```
## The bar_chart 

We know that the bar chart does this count automatically for us



```{r}
flights |> ggplot(aes(x = origin, fill = origin)) + geom_bar()
```

The above chart is same as

```{r}
flights |> count(origin) |> ggplot(aes(x = origin, y = n, fill = origin)) + geom_col()
```

## Inspecting group data

```{r}

carrier_grp <- group_by(flights, carrier)
nrow(flights) == nrow(carrier_grp)

length(colnames(flights))
length(colnames(carrier_grp))

group_vars(carrier_grp)
```

## Grouping by origin and dest


```{r}
flights |>
  group_by(origin, dest) |>
  group_vars()
```


```{r}
flights |> group_by(origin, dest) |>
  summarize(n = n())
```
## Relative frequency chart

Let us find the top 3 destinations in this flights dataset

```{r}
flights |> count(dest) |> slice_max(n = 3, n) -> top_3

top_3
```

Now let us find the relative frequencies - to show the top 3 destination distribution from the three origins.

```{r}

flights |> filter(dest %in% c(top_3$dest)) |> ggplot(aes(x = origin, fill = dest)) + geom_bar(position = 'fill')
```

## Exercise

Using the `mpg` data set, find the manufacturer (`manufacturer`) with the highest mean highway efficiency (`hwy`)

```{r manufacturer-hwy, exercise = TRUE}

```

```{r manufacturer-hwy-solution}
mpg |> group_by(manufacturer) |> summarize(mean_hwy = mean(hwy)) |> arrange(-mean_hwy)
```






## Dropping groups

After every summarize step, a `group var` is removed from the last group

Let us see an example

```{r}
flights |> group_by(origin, dest) |> summarize(n = n()) |> summarize(n = n())
```

What exactly did we get here? Let us analyze

```{r}
flights |> group_by(origin, dest) -> gp_data 

gp_data |> group_vars()

gp_data |> summarize(n = n()) |> group_vars()

```






## Useful functions for summaries

* Seen before: `mean`, `median`, `sd`, `min`, `max`
* Other common statistical measures: `quantile`, `IQR`
* For boolean/logical columns: `any` and `all` ("or" and "and" across vectors)
* The functions `n` and a `n_distinct` count units and distinct values

## Some more summaries

```{r}
flights |> group_by(carrier) |> summarize(n(), n_distinct(dest))
```

## Temperature data for Ann Arbor, MI

```{r}
aatemp
```

## `mutate` to add a column

```{r}
aatemp_cel <- mutate(aatemp,
                     TMAX_celsius = (TMAX - 32) * 5/9) |>
  select(TMAX, TMAX_celsius)

aatemp_cel |> head()
```

## Centered temperature

```{r}
mutate(aatemp, TMAX_centered = TMAX - mean(TMAX)) |>
 ggplot(aes(y = TMAX_centered, x = factor(quarter(DATE)))) +
    geom_violin() 
```

## `mutate` and `group_by`

Observe some care when using `mutate` on grouped tables:

```{r}
group_by(aatemp, quarter(DATE)) |>
  mutate(TMAX_centered = TMAX - mean(TMAX)) |>
  ggplot(aes(y = TMAX_centered, x = factor(`quarter(DATE)`))) +
    geom_violin()
```



## Normalizing by monthly averages?

Let's center each observation by it's monthly average that we can understand if
it was unusual for that time of year.

```{r}
aat_month_centered <- group_by(aatemp, month(DATE)) |>
  mutate(TMAX_centered = TMAX - mean(TMAX, na.rm = TRUE)) # mean computed over months
## verify 
summarize(aat_month_centered, var(TMAX_centered), sum(TMAX_centered^2) / (n() - 1)) |>
  head(3)
```

## Unusual months continued: conversion to ranks

*Ranks* are a useful robust replacement for values that are less susceptible to outliers. Let's rank days by how far they were from their monthly mean.

**Danger**: mutate will operate within months!

```{r}
mutate(aat_month_centered, r = rank(TMAX_centered)) |> 
  summarize(min(r), max(r))
```

## Ungrouping to fix

We need to drop the grouping values so that we can rank across all days.

```{r}
ungroup(aat_month_centered) |> 
  mutate(aat_month_centered, r = rank(TMAX_centered)) |>
  summarize(min(r), max(r))
```

## Average rank within years

Now that we can rank across all years and months, what year had the highest
average ranks?

```{r}
ungroup(aat_month_centered) |> 
  mutate(aat_month_centered, r = rank(TMAX_centered)) |>
  group_by(year(DATE)) |>
  summarize(mean(r)) |>
  arrange(desc(`mean(r)`))
```


## `rbind` and `cbind`

Take a sequence of vector, matrix or data-frame arguments and combine by rows you use `rbind`
And to combine them by columns you use `cbind`

```{r}
t1 <- tibble(x = 1, y = 2)
t1
t2 <- tibble(x = 2, y = 4)
t2
t <- rbind(t1, t2)
t
cbind(t, new_col = c('a', 'b'))
```

## Applying functions with `across`

You can define your own functions and use that function to apply to specific columns using `across`

Let us say you want to count the length of the string in every character column of the dataframe and create a new column with the string length value. You could do so as shown below:

```{r}
get_length <- function(x){
  str_length(x)
}

storms |> mutate(across(where(is.character), 
                        get_length, 
                        .names = 'length_{col}')) |> 
  head(3)
```


## Practice with everything

Let us look into the flights dataset and find the most delayed flight (highest departure delay) for each day of the year:

```{r}
library(nycflights13)
flights |>
  group_by(year, month, day) |>
    slice_max(dep_delay)
```

## 2nd most delayed?

```{r}
flights |> group_by(year, month, day) |>
  mutate(rank = min_rank(-dep_delay), .before = 'year') |>
    filter(rank == 2)
```

To do this question (easily) we used a new function, min_rank() (see Chapter 15 of the reading). This function takes a vector of numbers and returns their rank:

```{r}
min_rank(c(5, 1, 3, 10, 15))
```

## Most commonly delayed carrier

If I create a list of the ten most (departure) delayed flights for each day of the year, which carrier appears most frequently on that list?

```{r}
flights |> group_by(year, month, day) |>
  slice_max(n = 10, dep_delay) |> ungroup() |> count(carrier, sort = T)
```

## Proportion of delayed flights
We will call a flight delayed if dep_delay is greater an 10 minutes

```{r}
flights |>
  filter(!is.na(dep_delay)) |>
  group_by(carrier) |>
    summarise(
      total = n(),
      delayed = sum(dep_delay > 10),
      prop = delayed/total
    ) |> arrange(-prop)
```

