---
title: "DATASCI 306 Lab3"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

## Set up

- Use the install.packages("<PACKAGE NAME>") command to install packages in R
- To load a package (already installed) use the library(<PACKAGE NAME>) command

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# install.packages("tidyverse")
# install.packages("learnr")

```

```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)  # For ggplot2, dplyr
library(learnr)
library(dplyr)
```


## Table operations

Today we are going to focus on functions in the `dplyr` package, which is part of the `tidyverse`. 
The `dplyr` package aims to provide a function for each basic "verb" of data manipulation. 
Here are some common functions / "verbs" for data manipulation: 

* Rows: 
    + `filter()`: Select rows based on criteria on column values
    + `slice()`: choose rows based on locations.
    + `arrange()`: change the order of the rows.
* Columns: 
    + `select()`: Pick and choose columns to include in data.
    + `pull()`: Select a single column (similar to `data$column`).
    + `rename()`: Change the name of columns.
    + `mutate()`: Change the values of columns / creates new columns. 
* Groups (of rows):
    + `group_by()`: Creates a group of rows.
    + `summarise()` or `summarize()`: Calculate summary values for a given group. 

## Selecting, Filtering and Mutating

We are going to look at the `midwest` data again.

```{r}
head(midwest)
```

### `select`

If you might like to have the help file for the select function open, run

```{r eval = FALSE}
?select
```

in the Console tab.

### Exercise 1 {-}

#### (a) {-}

Use `select` to pick the columns that are related to population counts:

```
"poptotal"        "popdensity"      "popwhite"        "popblack"        "popamerindian"  
"popasian"        "popother"        "popadults"       "poppovertyknown"
```

```{r midwestSelect1, exercise=TRUE}

midwest %>%select(poptotal, popdensity, popwhite, popblack, popamerindian, popasian, popother, popadults, poppovertyknown)


```


#### (b) {-}

Select everything **except** `PID` and `poptotal`:

```{r midwestSelect2, exercise=TRUE}
midwest %>%
  select(-PID, -poptotal)

```

##### (c) {-}

Use the `where` and `is.character` function to select only columns containing strings/character vectors.

```{r midwestSelect3, exercise=TRUE}

midwest %>%
  select(where(is.character))

```



### `filter`
`filter()` allows us to select certain rows. 

Observe the following way to tell if a county is in Michigan:

```{r}
midwest$state == "MI"
```

#### Exercise 2 {-}

##### (a) {-}

Use code similar (but not the same as) the above to find all rows for counties in MI

```{r midwestFilter1, exercise=TRUE}

midwest %>%
  filter(state == "MI")

```


You can include more than one condition at a time by adding additional qualifications. 

```{r}
midwest %>%
  filter(state == "MI", county == 'WASHTENAW')
```

##### (b) {-}

Select all counties in Michigan with a `percollege` value greater than 17 (approximately the median in the entire data set).

```{r midwestFilter2, exercise=TRUE}

midwest %>%
  filter(state == "MI", percollege > 17)

```

Another way to do this filter is to use the "and" operator `&`

```{r}
midwest %>%
  filter(state == "MI" & percollege > 16)
```

There is also an "or" operator `|`.

##### (c) {-}

Use the `|` operator to find rows in either Wisconsin or Ohio.

```{r midwestOrOperator, exercise=TRUE}

midwest %>%
  filter(state == "WI" | state == "OH")

```

When filtering rows for many values in the same column, the `%in%` operator can be useful. Here's an example of it's usage

```{r}
c("WI", "MI", "OH", "OH", "IA") %in% c("WI", "OH")
```

##### (d) {-}

Use the `%in%` operator to repeat the previous selection of counties in Wisconsin or Ohio.

```{r midwestinoperator, exercise=TRUE}

midwest %>%
  filter(state %in% c("WI", "OH"))

```

### Putting `filter` and `select` together

##### (e) {-}

Make a table using all the rows from Wisconsin and Ohio but only the columns `PID` and `area`

```{r midwestfilterselect, exercise = TRUE}

midwest %>%
  filter(state %in% c("WI", "OH")) %>%
  select(PID, area)

```

## Mutate
`mutate()` can be used to either create or modify a column in a dataset. 
We can pipe multiple functions together. 

Try running the following code:
```{r mutate1, exercise=TRUE}
midwest %>%
  filter(state == 'MI') %>%
  mutate(school = ifelse(county == "WASHTENAW", "GO BLUE!!!", "OTHER")) %>%
  mutate(county == tolower(county)) %>%
  select(state, county, school) %>%
  tail()
```

#### Exercise 3 {-}

##### (a) {-}

Create a new column finds the number of non-adults per county (i.e., the difference between `poptotal` and `popadults`):

```{r mutate2, exercise=TRUE}
midwest %>%
  mutate(non_adults = poptotal - popadults) %>%
  select(state, county, poptotal, popadults, non_adults) %>%
  head()

```

##### (b) {-}

Mutate can make more than one column. Using only one call to `mutate`, create a `childrendensity` by dividing the number of children by area.

```{r mutate3, exercise=TRUE}

midwest %>%
  mutate(non_adults = poptotal - popadults,
         childrendensity = non_adults / area) %>%
  select(state, county, non_adults, area, childrendensity) %>%
  head()

```

We can use `arrange()` to order the columns of the data. 

```{r arrange1, exercise=TRUE}
midwest %>%
  select(state, county) %>%
  arrange(desc(state), county) %>%
  head()
```

##### (c) {-}

Find the county with the highest population (`poptotal`) in the data set:

```{r arrange2, exercise=TRUE}

midwest %>%
  arrange(desc(poptotal)) %>%
  select(state, county, poptotal) %>%
  head(1)

```


Another useful tool is `slice()`. This function allows us to select certain rows. There are a few versions of this that each give different results. See `?slice` for more details.

The previous example could have been written as:

```{r slice1, exercise=TRUE}
midwest %>%
  slice_max(poptotal)
```

## Grouping and Summarizing

`group_by()` helps us create groups. 
The following code will get the maximum population county for each state in the dataset. 

```{r groupby1, exercise=TRUE}
midwest %>%
  group_by(state) %>%
  slice_max(poptotal)
```

#### Exercise 4 {-}

##### (a) {-}

Use `group_by` and `slice_max` to find the largest metro and non-metro counties (`inmetro`):

```{r groupbymetro, exercise=TRUE}

midwest %>%
  group_by(inmetro) %>%
  slice_max(poptotal) %>%
  select(inmetro, state, county, poptotal)

```


`summarize()` (or equivalently, `summarise()`) is a function that's not particularly useful on it's own, but when combined with `group_by()` it can be very useful.
The following code counts the number of counties in each state

```{r summarize1, exercise=TRUE}
midwest %>%
  group_by(state) %>%
  summarize(n_counties = n()) %>%
  arrange(-n_counties)
```

##### (b) {-}

Use `group_by()` and `summarize()` to get the total population in each state (hint, we want to *sum* up the `poptotal` column). Then, arrange the states so that states with the smallest population comes first.

```{r groupExercise, exercise=TRUE}

midwest %>%
  group_by(state) %>%
  summarize(total_population = sum(poptotal)) %>%
  arrange(total_population)

```

##### (c) {-}

**BONUS** Modify your previous solution to also calculate the percentage of the population that is White, Black, and Asian. 

```{r groupExercisebonus, exercise=TRUE}

midwest %>%
  group_by(state) %>%
  summarize(total_population = sum(poptotal),
            pct_white = sum(popwhite) / total_population * 100,
            pct_black = sum(popblack) / total_population * 100,
            pct_asian = sum(popasian) / total_population * 100) %>%
  arrange(total_population)

```