---
title: "DATASCI 306: Lab 7"
author: "Your Name"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

## Set up

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.
# For HW4 Q2, Q3
if (!require(Lahman)) 
  install.packages("Lahman", repos = "http://cran.us.r-project.org")
```

```{r setup, eval=TRUE, include=FALSE, message=FALSE, echo=FALSE}
library(Lahman) # For HW4 Q2, Q3
library(learnr)     # For interactive exercises
library(lubridate)
library(tidyverse)  # For dplyr, stringr


tutorial_options(exercise.reveal_solution=TRUE)
```


## Today's Content

- Strings and Regular Expressions


## Strings

Recall that a **string** is a sequence of characters.

Below is an example of a string as it would appear on paper.
```
datascience
```

Strings in R work differently than they do on paper. In R, a string **must either begin and end with a double quote `"` or begin and end with a single quote `'`.**
```{r}
"datascience"
'datascience'
```

**Question:** What would happen if you entered `"datascience"` without the quotes and why?
```{r no_quotes, exercise=TRUE}

```

```{r no_quotes-solution}
datascience
# Without the quotes, the interpreter would assume that datascience is an
# object that we want to print. It wouldn't be able to find datascience because
# it doesn't exist and would then throw an error.
```

`writeLines()` takes the R version of a string and shows us the paper version.
```{r, comment=NA}
writeLines("datascience")
```

Certain characters are inserted in strings using **escape sequences**. An escape sequence starts with a backslash `\`. For example, the escape sequence for `tab` is `\t`. The backslash combines with the next character to form an escape sequence. See the help page on `Quotes` for the list of valid escape sequences in R. This is how a tab would be inserted in the previous string.
```{r}
"data\science"
```

This may be more natural, but it's worse because it's ambiguous.
```{r}
"data  science"
```

To include a backslash, use the escape sequence `\\`.
```{r}
"data\\science"
```

Another commonly used escape sequence is the newline escape sequence. A newline represents the start of a new line.
```{r}
"data\science"
```

**Question:** What would happen if you tried to insert a tab, or a backslash, or a newline in the middle of `"datascience"` without using an escape sequence and why? 
```{r escape_seqs, exercise=TRUE}

```

```{r escape_seqs-solution}
# Inserting a tab by hitting the tab key doesn't work. In the console, it would
# bring up the autocomplete box.
# Inserting a backslash with just one backslash causes \y to be interpreted as
# an escape sequence, which causes an error as \y isn't a valid escape
# sequence.
# Inserting a newline by hitting the return key seems to work, but R seems to
# think that the code has a mistake
```

Here are the outputs of `writeLines()` for the three modifications of `"datascience"`. They look the way they would on paper.
```{r, comment=NA}
writeLines("data\science")
writeLines("data\\science")
writeLines("data\science")
```

Since quotes are used to mark the start and end of an R string, an escape sequence must be used to put a quote in the middle of an R string if it's of the same type as the start and end quotes.
```{r, comment=NA}
"data\"science" 
writeLines("data\"science")
```

**Question:** What happens if you don't escape and why?
```{r no_escaping, exercise=TRUE}

```

```{r no_escaping-solution}
# The interpreter sees the middle quote and thinks that the string has ended,
# but then sees another character, which doesn't make any sense, so it throws
# an error
"data"science"
```

However, if the quote is of the other type, escaping isn't necessary.
```{r, comment=NA}
"data'science"
writeLines("data'science")
```

## Strings: More Functions

We've talked about several functions for working with strings: `str_c()`, `str_length()`, `str_sub()`, `str_split()`, etc. Some additional useful string functions are described below.

`strrep()` is used to repeat a string. `strrep()` is in the `base` package unlike the other string functions we've covered.
```{r}
strrep(c("apple", "banana", "cherry"), 1:3)
```

`str_interp()` makes a string from a template by inserting the values of variables. This can be very useful when making titles for plots. Note that a variable must be enclosed in braces and a dollar sign must be put before the left brace.
```{r}
x <- 1
y <- 2
z <- 3
str_interp("${x} + ${y} = ${z}")
```

### Exercise

Without hard-coding, recreate the ASCII art below. There are 10 lines, each of which has 8 hyphens in the middle.
```
|----------|
|----------|
|----------|
|----------|
|----------|
|----------|
|----------|
|----------|
|----------|
|----------|
```

```{r strrep, exercise=TRUE}

```

```{r strrep-hint}
# Try using strrep() and writeLines() in a for loop
```

```{r strrep-solution}
for (n in 1:10) {
  line <- str_c("|", strrep("-", 10), "|")
  writeLines(line)
}
```

### Exercise

Recall that `iris` is a data frame with measurements of petal and sepal length and width for a few flower species. `make_scatterplot()` takes a species and two measurement variables and makes a scatterplot of the variables for that species. Complete `make_scatterplot()`. If the species is `virginica`, the `x` variable is `Sepal.Length`, and the `y` variable is `Sepal.Width`, then the title of the plot should be `"Sepal.Width vs. Sepal.Length for virginica"`
```{r str_interp, exercise=TRUE}
make_scatterplot <- function(species, x_var, y_var) {
  title <- # Fill in
  iris %>%
    # Fill in - only use data for species
    ggplot(aes(!!as.name(x_var), !!as.name(y_var))) +
    # Fill in - choose the right geom
    # Fill in - set the title
    theme_bw()
}
make_scatterplot("virginica", "Sepal.Length", "Sepal.Width")
```

```{r str_interp-solution}
make_plot <- function(species, x_var, y_var) {
  title <- str_interp("${y_var} vs. ${x_var} for ${species}")
  iris %>%
    filter(Species == species) %>%
    ggplot(aes(!!as.name(x_var), !!as.name(y_var))) +
    geom_point() +
    ggtitle(title) +
    theme_bw()
}
make_plot("virginica", "Sepal.Length", "Sepal.Width")
```

### Optional Challenge Exercise

Without hard-coding, recreate the ASCII art below. There are 8 lines, each of which has 7 hyphens in the middle.
```
|\-------|
|-\------|
|--\-----|
|---\----|
|----\---|
|-----\--|
|------\-|
|-------\|
```

```{r challenge, exercise=TRUE}

```

```{r challenge-solution}
for (n in 0:7) {
  line <- str_c("|", strrep("-", n), "\\", strrep("-", 7 - n), "|") 
  writeLines(line)
}
```


## Regular Expressions

A **regular expression** is a string specifying a pattern that other strings may or may not match. Regular expressions can be used to

  - find strings matching a pattern
  - modify substrings matching a pattern
  - delete substrings matching a pattern
  
An example use of regular expressions is validation of email addresses. If a company wants to check whether a customer's email address is valid, they could check whether it matches a regular expression like [this one](https://stackoverflow.com/a/201378):

```
(?:[a-z0-9!#$%&'*+/=?^_`{|}~-]+(?:\.[a-z0-9!#$%&'*+/=?^_`{|}~-]+)*|"(?:[\x01-\x08\x0b\x0c\x0e-\x1f\x21\x23-\x5b\x5d-\x7f]|\\[\x01-\x09\x0b\x0c\x0e-\x7f])*")@(?:(?:[a-z0-9](?:[a-z0-9-]*[a-z0-9])?\.)+[a-z0-9](?:[a-z0-9-]*[a-z0-9])?|\[(?:(?:(2(5[0-5]|[0-4][0-9])|1[0-9][0-9]|[1-9]?[0-9]))\.){3}(?:(2(5[0-5]|[0-4][0-9])|1[0-9][0-9]|[1-9]?[0-9])|[a-z0-9-]*[a-z0-9]:(?:[\x01-\x08\x0b\x0c\x0e-\x1f\x21-\x5a\x53-\x7f]|\\[\x01-\x09\x0b\x0c\x0e-\x7f])+)\])
```

### Metacharacters

Several characters have special meanings inside regular expressions. They are called **metacharacters**. These are the metacharacters:

`. \ | ( ) [ ] ^ $ { } * + ?`

For example, `.` matches any character.

If you do not want a metacharacter to have its special meaning, you need to prepend it with **two** backslashes.
```{r}
str_view_all(c("$2.50", "2 dollars"), "\\.")
```

**Question:** What happens if you use only one backslash or no backslashes and why?
```{r too_few_backslashes, exercise=TRUE}

```

```{r too_few_backslashes-solution}
# Both strings match since . matches any character
str_view_all(c("$2.50", "2 dollars"), ".")
# Causes an error since \. isn't a valid escape sequence
str_view_all(c("$2.50", "2 dollars"), "\.")
```

An alternative is to use `fixed()`.
```{r}
str_view_all(c("$2.50", "2 dollars"), fixed("."))
```

### Character Classes

- `.` matches any character
- `[abcd]` matches any one of the characters between the brackets
- `[^abcd]` matches any character *not* between the brackets (excluding `^`)
- `[a-d]` matches any character in the specified range; it's the same as `[abcd]`

An example:
```{r}
str_view_all(c("$2.50", "2 dollars"), "[a-zA-Z]")
```

An alternative way to make the matching case-insensitive:
```{r}
str_view_all(c("$2.50", "2 dollars"), regex("[a-z]", ignore_case = TRUE))
```


### Shorthand

- `\w` matches a "word" character, equivalent to `[0-9A-Za-z_]`, i.e., digits, letters, and the underscore
- `\s` matches any whitespace character, including tabs and newlines
- `\d` matches digits, equivalent to `[0-9]`
- `\W`, `\S`, and `\D` match the opposite of the lower-case versions

In the example below, we check which string contains whitespace. Note that **two** backslashes need to be used. Why?
```{r}
str_view_all(c("$2.50", "2 dollars"), "\\s")
```

### Grouping

- `()` are used to group patterns together. This can be used to extract matches individually.
- `\1` refers to the match to the first group, `\2` refers to the match to the second group, etc.

In this example, we check which string contains a repeated letter.
```{r}
str_view_all(c("$2.50", "2 dollars"), "([a-z])\\1")
```

### Operators

- `|` is the OR operator and allows matches of either side
- `{}` describes how many times the preceeding character or group must occur:
  - `{m}` must occur exactly `m` times
  - `{m,n}` must occur between `m` and `n` times, inclusive
  - `{m,}` Must occur at least `m` times
- `*` means the preceeding character can appear zero or more times, equivalent to `{0,}`
- `+` means the preceeding character must appear one or more times, equivalent to `{1,}`
- `?` means the preceeding character can appear zero or one time, equivalent to `{0,1}`

Which string has two digits in a row?
```{r}
str_view_all(c("$2.50", "2 dollars"), "[0-9]{2}")
```

### Anchors

- `^` matches the start of a string (or line)
- `$` matches the end of a string (or line)
- `\b` matches a word boundary
- `\B` matches not word boundary

Which string starts with a dollar sign? Note that `$` is escaped with **two** backslashes.
```{r}
str_view_all(c("$2.50", "2 dollars"), "^\\$")
```


## Regular Expression Exercises

The exercises below use this paragraph.

```{r}
baseball <- "According to Baseball Reference’s wins above average, The Red Sox had the best 
outfield in baseball— one-tenth of a win ahead of the Milwaukee Brewers, 11.5 to 11.4. And 
that’s despite, I’d argue, the two best position players in the NL this year (Christian 
Yelich and Lorenzo Cain) being Brewers outfielders. More importantly, the distance from 
Boston and Milwaukee to the third-place Yankees is about five wins. Two-thirds of the Los 
Angeles Angels’ outfield is Mike Trout (the best player in baseball) and Justin Upton (a 
four-time All-Star who hit 30 home runs and posted a 122 OPS+ and .348 wOba this year), 
and in order to get to 11.5 WAA, the Angels’ outfield would have had to replace right 
fielder Kole Calhoun with one of the three best outfielders in baseball this year by WAA."
```

The exercises require you to extract matches using `str_extract_all()`

### Exercise 1

Extract all the capitalized words in the paragraph.

```{r baseball1, exercise=TRUE}

```

```{r baseball1-hint}
# Try using \b
```

```{r baseball1-solution}
# `\b` looks for a word boundary (not just the beginning of the text snippet!)
# `[A-Z]` matches a single capitalized letter
# `[a-z]` matches a single lowercase letter
# `+` means we match arbitrarily many lowercase letters
str_extract_all(baseball, "\\b[A-Z][a-z]+")
```

### Exercise 2

Extract all the numbers.

```{r baseball2, exercise=TRUE}

```

```{r baseball2-solution}
str_extract_all(baseball, "\\.?\\d+\\.?\\d*")
```

### Exercise 3

Extract all the hyphenated words.

```{r baseball3, exercise=TRUE}

```

```{r baseball3-solution}
str_extract_all(baseball, regex("[a-z]+-[a-z]+", ignore_case = TRUE))
```

### Exercise 4

Extract all words with two consecutive vowels.

```{r baseball4, exercise=TRUE}

```

```{r baseball4-solution}
str_extract_all(baseball, regex("[a-z]*[aeiou]{2}[a-z]*", ignore_case = TRUE))
```

### Exercise 5

Extract all words with a repeated letter.

```{r baseball5, exercise=TRUE}

```

```{r baseball5-solution}
str_extract_all(baseball, regex("[a-z]*([a-z])\\1[a-z]*", ignore_case = TRUE))
```
