---
title: "DATASCI 306 Lab2"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Plotting with `ggplot2`

Recall from lecture that a plot made using `ggplot2` has several components:

* The graph object itself (creating using `ggplot(data)`)
* A set of *aesthetic* mappings (connecting data to visual variables)
* Layers: collections of geometric elements (`geom_*()`) and statistical transformations (`stat_*()`)
* Scales: information on the range or composition of variables
* Coordinate systems: how the data are arranged spatially
* Facet: breaking a single plot into many, similar plots
* Theme: all the other color and printing aspects of the plot

Today we'll talk about aesthetic mappings, geometric elements, and statistical transformations.

## Aesthetic Mappings

An *aesthetic* is a visual property of the geometric elements in a plot.

```{r aesthetic_mappings_example1}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, color = drv))
```

Geometric elements represent observations in the dataset. In the plot above, the points are the geometric elements. Different geometric elements have different aesthetics. Some aesthetics of points are

* `x`, or x-coordinate
* `y`, or y-coordinate
* `color`
* `shape`
* `size`

The background, the gridlines, and the tick marks are not geometric elements because they don't represent any observations. This means that properties like the background color aren't aesthetics.

An *aesthetic mapping* is an assignment of an aesthetic to a variable in the dataset. Aesthetic mappings are listed inside `aes()`. For example, in the previous plot, the aesthetic `x` was mapped to the variable `cty`. 

Each layer has a set of aesthetic mappings. By default, a layer inherits its mappings from the base created by `ggplot()`. So, the code for the plot above could be changed to

```{r aesthetic_mappings_example2}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy, color = drv)) + 
  geom_point()
```

Mapping `color` to a categorical variable will cause the observations to be grouped by category in addition to being colored by category. That's why we get one boxplot for each drive train or `drv` category in the plot below.

```{r mapping_color_example}
ggplot(data = mpg, mapping = aes(y = hwy, color = drv)) + 
  geom_boxplot()
```

To explicitly group by drive train, we can map the `group` aesthetic to `drv`. However, if we don't also map `color`, then the boxplots won't be colored differently, and there won't be a legend explaining which boxplot is for which drive train.

```{r mapping_group_example}
ggplot(data = mpg, mapping = aes(y = hwy, group = drv)) + 
  geom_boxplot()
```

### Exercise 1 {-}

#### (a) {-}

Modify the code below so that the point shape varies by drive train.

```{r exercise1a, exercise=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, color = drv))
```

```{r exercise1a-solution}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, color = drv, shape = drv))
```

#### (b) {-}

Modify the code below so that each point is a triangle. How is your code different from your code for part (a)? Why is it different?

*Hint: use the string "triangle"*

```{r exercise1b, exercise=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, color = drv))
```

```{r exercise1b-solution}
# The key difference here is that the shape = part goes outside aes() instead
# of inside aes(). This is because here we're setting shape to a constant
# instead of mapping it to a variable. Only aesthetic mappings go inside
# aes().
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cty, y = hwy, color = drv), shape = "triangle")
```

## Geometric Elements

A *geometric element* or *geom* is a geometric object that is used in a plot to represent observations. A geom can represent

  * one observation;
  * a group of observations;
  * all of the observations.

A `ggplot2` plot consists of a base created by `ggplot()` and zero or more layers, each representing observations in a particular dataset using a particular geom. There are many geoms; several commonly used geoms are described below. The examples visualize a dataset called `midwest` that has data from the 2000 census on counties in several midwestern states.

```{r convert_inmetro}
# inmetro is an integer that equals 0 or 1; plots will look better if it's
# logical
midwest <- mutate(midwest, inmetro = as.logical(inmetro))
```

### Plotting the Distribution of One Variable

If we want to visualize the distribution of a single variable, we first need to determine whether it's categorical or numerical.

If the variable is categorical, then we can make a bar chart. We use `geom_bar()` to do this. The plot below shows the number of counties by state.

```{r geom_bar_example}
midwest %>%
  ggplot(aes(state)) +
  geom_bar()
```

There are several options if the variable is numerical. `perchsd` gives the percentage of high school graduates. We could make a histogram of `perchsd` using `geom_histogram()`.

```{r geom_histogram_example}
midwest %>%
  ggplot(aes(perchsd)) +
  geom_histogram()  
```

We could also make a frequency polygon with `geom_freqpoly()`.

```{r geom_freqpoly_example}
midwest %>%
  ggplot(aes(perchsd)) +
  geom_freqpoly()
```

Finally, we could use `geom_boxplot()` to make a boxplot.

```{r geom_boxplot_example1}
midwest %>%
  ggplot(aes(perchsd)) +
  geom_boxplot()
```

### Plotting the Relationship between Two Variables

Just like in the one variable case, when we have two variables, we first need to determine the types of the variables before we make our plot.

We can make a scatterplot with `geom_point()` if both variables are continuous variable. The plot below displays the relationship between `perchsd` and `percadultpoverty`, the percentage of adults below the poverty line.

```{r geom_point_example}
midwest %>%
  ggplot(aes(perchsd, percadultpoverty)) +
  geom_point()
```

Instead of plotting all of the points, we could plot a *smooth*, a curve that summarizes the trend in the data. This is done with `geom_smooth()`.

```{r geom_smooth_example}
midwest %>%
  ggplot(aes(perchsd, percadultpoverty)) +
  geom_smooth(se = FALSE)
```

#### Exercise 2 {-}

##### (a) {-}

Make a plot of `percadultpoverty` versus `perchsd` that has both points and a smooth.

```{r exercise2a, exercise=TRUE}

```

```{r exercise2a-solution}
midwest %>%
  ggplot(aes(perchsd, percadultpoverty)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

##### (b) {-}

Make a plot that is the same as the one in part (a), except with a confidence interval around the smooth. What do you notice about the width of the interval? Why does it make sense for the width to be how it is?

*Hint: look at the help page for `geom_smooth()` to make the interval*

```{r exercise2b, exercise=TRUE}

```

```{r exercise2b-solution}
# The width increases toward the sides of the plot. This makes sense, as there
# are few points near the sides, so the estimates of the trend there should be
# highly uncertain.
midwest %>%
  ggplot(aes(perchsd, percadultpoverty)) +
  geom_point() +
  geom_smooth() # Inserting se = TRUE is unnecessary since se = TRUE by default
```
##### (c) {-}

You might have noticed that the smooth in part (b) was calculated using a method called LOESS. Other methods can be used as well. Remake the plot, but with a smooth calculated using linear regression.

*Hint: look at the help page for `geom_smooth()` again*

```{r exercise2c, exercise=TRUE}

```

```{r exercise2c-solution}
midwest %>%
  ggplot(aes(perchsd, percadultpoverty)) +
  geom_point() +
  geom_smooth(method = "lm") # lm() is the function used for linear regression
```

`box and whisker plot` is meant for a continuous variable. When you want to group records based on a categorical variable and then draw a boxplot for the subset of the grouped data, you can use both the categorical and continuous as shown below.

```{r geom_boxplot_example2}
midwest %>%
  ggplot(aes(state, percollege)) +
  geom_boxplot()
```

We could also make a bar chart using `geom_col()`. Below, `midwest2` contains one row per state in `midwest` and has the number of counties for each state. We can use `geom_col()` to recreate the bar chart made with `geom_bar()` above.

```{r geom_col_example}
midwest2 <- midwest %>%
  group_by(state) %>%
  summarize(num_counties = n())
midwest2 %>%
  ggplot(aes(state, num_counties)) +
  geom_col()
```

#### Exercise 3 {-}

##### (a) {-}

A `histogram` is another plot meant for a continuous variable.
Instead of drawing multiple boxplots to visualize the relationship between `percollege` and `state`, we could use histograms. Make one plot that has a histogram of `percollege` for each value of `state`. Give the histograms different colors. How does this plot compare to the boxplot?

```{r exercise3a, exercise=TRUE}

```

```{r exercise3a-solution}
# This plot is harder to read than the boxplot. However, unlike the boxplot, it
# does show that the different states have different numbers of counties.
midwest %>%
  ggplot(aes(percollege, color = state)) +
  geom_histogram()
```

##### (b) {-}

Now make a one plot that has a frequency polygon for each state instead of a histogram. Give the frequency polygons different colors. How does this plot compare to the preceding plots?

```{r exercise3b, exercise=TRUE}

```

```{r exercise3b-solution}
# This plot is also hard to read, and it also shows the count information that
# the boxplot doesn't show.
midwest %>%
  ggplot(aes(percollege, color = state)) +
  geom_freqpoly()
```
##### (c) {-}

Complete the code below to use boxplots to visualize the relationship between `inmetro` and `percollege`. `inmetro` is `TRUE` if a county is in a metro area and false otherwise.

```{r exercise3c, exercise=TRUE}
midwest %>%
  mutate(inmetro = as.logical(inmetro)) %>%
  ggplot()
```

```{r exercise3c-solution}
midwest %>%
  mutate(inmetro = as.logical(inmetro)) %>%
  ggplot(aes(inmetro, percollege)) +
  geom_boxplot()
```

##### (d) {-}

Now complete the code to use frequency polygons to visualize the relationship. Is this plot worse than the boxplot?

```{r exercise3d, exercise=TRUE}
midwest %>%
  mutate(inmetro = as.logical(inmetro)) %>%
  ggplot
```

```{r exercise3d-solution}
# Unlike the previous frequency polygon plot, this one isn't hard to read. This
# plot provides count information that the boxplot doesn't provide. So, unlike
# previously, the frequency polygon plot isn't worse than the boxplot.
midwest %>%
  mutate(inmetro = as.logical(inmetro)) %>%
  ggplot(aes(percollege, color = inmetro)) +
  geom_freqpoly()
```

Finally, if both variables are categorical, we can plot them using `geom_count()`.

```{r geom_count_example}
midwest %>%
  ggplot(aes(state, inmetro)) +
  geom_count()
```

## Statistical Transformations

A *statistical transformation* or *stat* is an algorithm that computes from raw data values the values that will be plotted. The available stats are implemented in functions whose names begin with `stat_`. A stat is applied even when the raw values themselves are to be plotted. In that case, the stat that is applied is `stat_identity()`. For example, suppose that our dataset is

```{r stat_identity_example1}
(tbl <- tibble(x = 2^(-3:3), y = log(x)))
```

The result of applying `stat_identity()` to `tbl` would be `tbl`.

Each geom has a default stat and each stat has a default geom. We can see which stat is the default stat for a geom by looking at the default value of the geom's `stat` argument. For instance, the default value of `stat` for `geom_point()` is `"identity"`.

```{r geom_point_default_stat}
geom_point
```

In a similar way, we can view the default geom for a stat:

```{r stat_identity_default_geom}
stat_identity
```

Because of the above, geoms and stats are interchangeable, i.e., you can do `ggplot() + stat_*()` instead of `ggplot() + geom_*()`. Thus, to make a scatterplot of `y` versus `x`, we could use the code below:

```{r geom_point_example2}
ggplot(tbl, aes(x, y)) +
  geom_point()
```

Or we could use the code below:

```{r stat_identity_example2}
ggplot(tbl, aes(x, y)) +
  stat_identity()
```

Stats are more of an issue for bar charts than for scatterplots. The default stat for `geom_bar()` is `stat_count()`. Suppose we wanted to use `midwest` to count counties in midwestern states like we did in a previous section. `stat_count()` takes `midwest` and computes the below tibble from it:

```{r stat_count_example}
(midwest2 <- midwest %>%
  group_by(state) %>%
  summarize(count = n(), prop = n() / n()) %>%
  ungroup())
```

`geom_bar()` uses the new tibble to construct the plot - `x` is mapped to `state` and `y` is mapped to `count`.

```{r geom_bar_example2}
ggplot(midwest, aes(state)) +
  geom_bar()
```

If we wanted to make the mapping of `y` explicit, we could use `after_stat()` as below. `after_stat(count)` indicates that we want to map the `count` variable in the tibble created by the stat.

```{r after_stat_example1}
ggplot(midwest, aes(state, after_stat(count))) +
  geom_bar()
```

Note that a `prop` variable was also created by the stat. We can use `after_stat()` to map `prop` instead of `count`. This causes the height of each bar to be a proportion instead of a count. `group = 1` ensures that the denominator of each proportion is the total number of counties.

```{r after_stat_example2}
ggplot(midwest, aes(state, after_stat(prop), group = 1)) +
  geom_bar()
```

### Exercise 4 {-}

#### (a) {-}

Let's return to the situation considered in Exercise 2(a). Again make a plot of `percadultpoverty` versus `perchsd` that has both points and a smooth, but change a setting to make the smooth much wigglier. Why is it a bad idea to do this?

*Hint: look at an appropriate help page*

```{r exercise4a, exercise=TRUE}

```

```{r exercise4a-solution}
# A very wiggly smooth might capture quirks of the data that are due to chance
# variation instead of the underlying pattern
midwest %>%
  ggplot(aes(perchsd, percadultpoverty)) +
  geom_point() +
  stat_smooth(span = 0.1) # Any small value is acceptable here
```

#### (b) {-}

Like you did in Exercise 2(b), make a scatterplot of `percadultpoverty` versus `perchsd` with a confidence interval around the smooth. This time, create a confidence interval whose level is 0.5. Why is the interval different from the previous interval?

*Hint: take another look at the documentation*

```{r exercise4b, exercise=TRUE}

```

```{r exercise4b-solution}
# The confidence level for the previous interval was 0.95. The smaller the
# confidence level, the narrower the interval is.
midwest %>%
  ggplot(aes(perchsd, percadultpoverty)) +
  geom_point() +
  stat_smooth(level = 0.5)
```