---
title: "DataSci 306: Lab 7"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

## Set up

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.

if (!require(nycflights13)) install.packages("nycflights13", repos = "http://cran.us.r-project.org")
if (!packageVersion("learnr") == "0.11.1") install.packages("learnr", repos = "http://cran.us.r-project.org")
```

```{r setup, eval=TRUE, include=FALSE, message=FALSE, echo=FALSE}
library(learnr)     # For interactive exercises
library(tidyverse)  # For ggplot2, dplyr, readr
library(lubridate)  # For some functions involving dates
library(nycflights13)
library(readxl)

tutorial_options(exercise.reveal_solution = FALSE)
```

```{r fake_setup, echo=TRUE, eval=FALSE}
library(learnr)     # For interactive exercises
library(tidyverse)  # For ggplot2, dplyr, readr
library(lubridate)  # For some functions involving dates
library(nycflights13)
library(readxl)
```


## Today's Content

* Data files handling

## Spreadsheets
In this section, we will learn how to import spreedsheets, which is covered in Chapter 22 of R4DS

In Chapter 8 Data import, we have learned how about importing data from plain txt files, e.g. .csv and .tsv files. However, we may also need to analyze data from spreadsheets sometimes. We will introduce tools for importing and writing data data from Excel spreadsheets and Google Sheets.

* Excel
* Reading spreadsheets
* Writing to Excel

## Reading Spreadsheets
here are some example excel sheets in `readxl` package, you can find the paths by calling `readxl_example()`

```{r}
readxl_example() # some example excel sheets in readxl package
geometry_path = readxl_example("geometry.xlsx")
geometry_path # show location of the dataset
```

## Read excel

```{r}
read_excel(geometry_path)
# equivalent to
# read_excel('/usr/local/lib/R/site-library/readxl/extdata/geometry.xlsx')
```

You can change column name while reading the Excel file.

```{r}
read_excel(geometry_path, col_names = c("A", "B", "C"))
```

```{r}
deaths_path = readxl_example("deaths.xlsx")
head(read_excel(deaths_path))
```

## using head

The output says there are 5 columns unnamed and named automatically.

```{r}
head(read_excel(deaths_path, 
    col_names = c("Name", "Profession", "Age", "Has kids", "DOB", "DOD")))
```

If there are multiple worksheets, you can read a single worksheet from a spreadsheet with `sheet` arguemnt. \ There are 4 worksheets in `datasets.xlsx: iris, mtcars, chickwts`, and `quakes`

```{r}
head(read_excel(readxl_example('datasets.xlsx')))
head(read_excel(readxl_example('datasets.xlsx'), sheet = 'iris'))
head(read_excel(readxl_example('datasets.xlsx'), sheet = 'mtcars'))
```

The `death` dataset looks messy. Let's clean it up! \ Let's first take a look at the spreadsheet view of `geom_death.xlsx`.

## Other functions
In read_excel function,

* skip = k will skip the first k rows.
* n_max = n will take at most n rows.
* range = "A6:C9" will read cells from A6 to C9.
* range = cell_rows(c(a, b)) will read rows from a to b.
* range = cell_cols(c(a, b)) will read columns from a to b.

```{r}
read_excel(deaths_path, skip = 4, n_max = 10) # skip the first 4 rows and the fifth row turns into column names
read_excel(deaths_path, range = "A6:C9")
read_excel(deaths_path, range = cell_cols(c(2, 3))) # omit NA at first rows
```

You may have noticed that, there are different data types in the spreadsheet. In csv file, all values are string. In Excel files, a cell can be one of five things:

* A boolean, like TRUE, FALSE, or NA
* A number, like “10” or “10.5”
* A datetime, which can also include time like “11/1/21” or “11/1/21 3:00 PM”
* A text string, like “ten”

## Other reader methods

he readr package in tidyverse is used for importing datasets:

* read_csv(): comma-separated values (CSV)
* read_tsv(): tab-separated values (TSV)
* read_csv2(): semicolon-separated values with , as the decimal mark
* read_delim(): delimited files (CSV and TSV are important special cases)
* read_fwf(): fixed-width files
* read_table(): whitespace-separated files
* read_log(): web log files


## Writing to Excel

```{r}
# install.packages('writexl')
library(writexl)
```


```{r}

deaths = read_excel(deaths_path, skip = 4, n_max = 10)
write_xlsx(deaths, path="deaths_cleaned.xlsx")
```

```{r}
out <- read_xlsx("deaths_cleaned.xlsx")
all.equal(out, deaths)
```

## Google Sheets

```{r}
# install.packages("googlesheets4")
library(googlesheets4)
gs4_deauth() # Call `gs4_deauth()` to prevent the attempt to get credentials.
# gs4_auth(email = "mine@example.com") # get credentials to access restricted file. 
```

`read_sheet()` reads a Google Sheet from a URL or file id. This function is equivalent to `range_read()`

```{r}

penguins <- read_sheet("https://docs.google.com/spreadsheets/d/1aFu8lnD_g0yjF5O-K6SFgSEWiHPpgvFCF0NY9D6LXnY/edit?usp=sharing")
penguins2 <- read_sheet("1aFu8lnD_g0yjF5O-K6SFgSEWiHPpgvFCF0NY9D6LXnY")
all.equal(penguins, penguins2)
head(penguins)
```
To get a quick view of the meta data of the sheet, use gs4_get()

```{r}
gs4_get("1aFu8lnD_g0yjF5O-K6SFgSEWiHPpgvFCF0NY9D6LXnY")
```

You can obtain a list of all sheets within a Google Sheet with sheet_names().

```{r}
sheet_names("1aFu8lnD_g0yjF5O-K6SFgSEWiHPpgvFCF0NY9D6LXnY")
```

Use `sheet` argument in read_sheet() to read different worksheets.

```{r}
head(read_sheet("1aFu8lnD_g0yjF5O-K6SFgSEWiHPpgvFCF0NY9D6LXnY", sheet="Biscoe Island"))
```

## Exercise 1
Read the Google Sheet titled survey from link, with survey_id as a character variable and n_pets as a numerical variable.

Hint: Column Specification

```{r q1, exercise = TRUE}

```


```{r q1-solution}
gs4_get("1yc5gL-a2OOBr8M7B3IsDNX5uR17vBHOyWZq6xSTG2G8")
read_sheet("1yc5gL-a2OOBr8M7B3IsDNX5uR17vBHOyWZq6xSTG2G8", col_types = "cd")
```

## Exercise 2
Read the first 6 rows (including column names) from third to fifth columns in worksheet `Dream Island` in Google Sheet file penguins using `read_sheet()` with file_id: 1aFu8lnD_g0yjF5O-K6SFgSEWiHPpgvFCF0NY9D6LXnY.

```{r q2, exercise = TRUE}

```

```{r q2-solution}
read_sheet("1aFu8lnD_g0yjF5O-K6SFgSEWiHPpgvFCF0NY9D6LXnY", range="Dream Island!C1:E6")
```
