---
title: 'DATASCI 306: Lab 2b'
author: "Your Name"
output: html_document
---

## Set up

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.

if (!require(ggforce)) install.packages("ggforce", repos = "http://cran.us.r-project.org")
if (!require(maps)) install.packages("maps", repos = "http://cran.us.r-project.org")
if (!require(mapproj)) install.packages("mapproj", repos = "http://cran.us.r-project.org")
if (!require(lattice)) install.packages("lattice", repos = "http://cran.us.r-project.org")
```

```{r setup, echo=TRUE}
library(tidyverse)  # For ggplot2, dplyr
library(maps)       # For plotting maps 
library(ggforce)    # For hurricane map, later. 
```

## Facets 

Faceting allows us to split data into subset and plot similar graphs for each subset. 
In lecture we saw that facets are often very useful to avoid over-fitting. 
There are three types of facets in `ggplot2`: 

* `facet_wrap`: Creates a one-dimensional set of plots, that are often shown in a 2d grid.
* `facet_grid`: Creates a two-dimensional grid of panels, defined by two different columns. 
* `facet_null`: This is the default facet, which is just a single plot

## Difference between `facet_wrap` and `facet_grid`

\
\
\

<center>

![A sketch difference between facetting systems. Figure borrowed from [ggplot2 book](https://ggplot2-book.org/facet.html)](images/facets.png)

</center>

## `facet_wrap` Example

Lets use the `ToothGrowth` dataset:

```{r headTooth, exercise=TRUE}
head(ToothGrowth)
```

```{r uniqueDose, exercise=TRUE}
unique(ToothGrowth$dose)
```

A box-plot of all of the data: 

```{r}
tooth_gg <- ggplot(ToothGrowth) + 
  geom_boxplot(aes(y = len)) + 
  theme_bw()

print(tooth_gg)
```


```{r createBasicBox, echo=TRUE, include=FALSE}
tooth_gg <- ggplot(ToothGrowth) + 
  geom_boxplot(aes(y = len)) + 
  theme_bw()
```

Now if we want to summarize at the distribution for tooth growth for each supplement, we can facet on supplement: 

```{r wrappedBox, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_wrap(~supp)
```

If we wanted to have two rows of figures instead of two columns, we can specify this with `ncol = 1` or `nrow = 2`

```{r wrappedBox2, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_wrap(~supp, ncol = 1)
```

We can get the same figure (but with labels on the side) by using `facet_grid`.

```{r gridBox1, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_grid(supp~.)
```

In the previous code, we are saying that we want the rows of the grid to correspond to the variable `supp`, and we want to give it an empty variable `.` for the columns. 
We can "rotate" the figure by switching the order of these variables: 

```{r gridBox2, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_grid(.~supp)
```

We can of course include a second variable rather than a missing variable if we wanted to do so: 

```{r gridBox3, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_grid(supp~dose)
```

The order of the variables depends on what information we are most interested in showing.
For example, in the previous plot, the order makes it easiest to compare the effect of various different levels of dose for each supplement. 
If we were interested how supplements compare for similar dosages, the following plot may be preferred:

```{r gridBox4, exercise=TRUE, exercise.setup="createBasicBox"}
tooth_gg + 
  facet_grid(dose~supp)
```

## `facet_wrap` Exercise

For this exercise, we are going to use the `barley` data from the `lattice` package. 
This is famous statistical dataset that has been cited in many papers.
We will use it to create a *Cleveland dot plot*. 

```{r}
barley <- lattice::barley
head(barley)
```

```{r loadBarley}
barley <- lattice::barley
```

**Exercise**: Add a facets and colors to the following code to reduce over-plotting. 
Do you notice anything strange about the data? 

```{r facetExercise, exercise=TRUE, exercise.setup="loadBarley"}
ggplot(barley) + 
  geom_point(aes(x = yield, y = variety)) + 
  theme_bw()
```

## `facet_wrap` Solution 

William Cleveland (to whom the term "data science" is often attributed) famously made a plot similar to the following in a textbook he published:

```{r}
ggplot(barley) + 
  geom_point(aes(x = yield, y = variety, col = year)) + 
  facet_wrap(~site) + 
  theme_bw()
```

In this plot, he points out that an oddity in Morris: it's the only site in which the 1932 crop yield was higher than that in 1931! 
Using this plot, he claimed that the data must have been mislabeled (although there is [some recent argument](https://blog.revolutionanalytics.com/2014/07/theres-no-mistake-in-the-barley-data.html) that this is not the case). 

## Coordinate Systems

In our last lecture, we looked closely at various *coordinate systems* in `ggplot2`. 
Coordinate systems allow us to easily change how the data are displayed, with limited effort. 

There are *linear* coordinate systems that don't change the shape of `geoms`: 

> - `coord_cartesian()`: Default coordinate system
> - `coord_flip()`: Default coordinate system, but flipping `x` and `y` axis. 
> - `coord_fixed()`: Default coordinate system, but fixing the ratio between `x` and `y` axis. 

There are also *non-linear* coordinate systems, that can change the shape of `geoms`: 

> - `coord_polar()`: Polar coordinates.
> - `coord_map()`: Map projections (spherical earth to 2D plane).
> - `coord_trans()`: Custom transformations to `x` and `y` positions. 

## Polar Coordinate Example

One of the most common coordinate transformations we see for data visualizations is **Polar-Coordinates**. 
The idea for polar coordinate mapping is that sometimes displaying the information as an angle and radius is more informative than displaying the information in the `x, y` Cartesian plane,
for example polar coordinates often help us visualize seasonal effects. 

```{r mtcarsPolar, exercise=TRUE}
my_cars <- mtcars
my_cars$model <- rownames(mtcars)
ggplot(my_cars) + 
  geom_bar(aes(y = mpg, x = model), stat = 'identity') + 
  coord_polar() + 
  theme_bw()
```

## `coord_map()` Example

Let's take a look at a new coordinate system, using `coord_map()`. 
First, we will use the `maps` package to get information about some North American Countries. 

```{r NorthAmerica, echo=TRUE}
world <- map_data("world") %>%
  filter(region %in% c("USA", "Canada", "Mexico", "Jamaica",
                       "Cuba", "Panama", "Guatemala", "Puerto Rico", 
                       "Costa Rica", "Dominican Republic", "Haiti",
                       "El Salvador", "Belize", "Bahamas", "Nicaragua",
                       "Honduras", "Trinidad")) %>%
  filter(long < 0)
```

We can plot this data using `ggplot2` and `geom_polygon()`. 
Lets save the map to use later:

```{r plotNorthAmerica, echo=TRUE}
my_map <- ggplot(world) + 
  geom_polygon(aes(x = long, y = lat, group = group), fill = 'white',
               col = 'black') + 
  theme_bw()

print(my_map)
```

## `coord_map()` continued

```{r preLoadMap}
world <- map_data("world") %>%
  filter(region %in% c("USA", "Canada", "Mexico", "Jamaica",
                       "Cuba", "Panama", "Guatemala", "Puerto Rico", 
                       "Costa Rica", "Dominican Republic", "Haiti",
                       "El Salvador", "Belize", "Bahamas", "Nicaragua",
                       "Honduras", "Trinidad")) %>%
  filter(long < 0)

my_map <- ggplot(world) + 
  geom_polygon(aes(x = long, y = lat, group = group), fill = 'white',
               col = 'black') + 
  theme_bw()
```

The problem with the previous slide is that the polygons are a bit distorted, so the polygons don't look exactly like we would expect.
`coord_map()` can help fix this issue.
In particular, we will start with `coord_quickmap()`, which is the simplest implementation. 

```{r coordQuickMap, exercise=TRUE, exercise.setup="preLoadMap"}
my_map + 
  coord_quickmap()
```

## `coord_map()` Continued

There are probably better ways to project the 3D world onto a 2D surface. 
You can explore some of these options by looking at the help page for `coord_map` and `mapproj::mapproject`.
Let's use the [globular](https://en.wikipedia.org/wiki/Nicolosi_globular_projection) projection

```{r globularMap, exercise=TRUE, exercise.setup="preLoadMap"}
my_map + 
  coord_map('globular')
```

## Hurricane Map

For this next plot, we will use the built in `storms` dataset, and we will look particularly at the storms in $2016$. 

```{r HurricaneMap, exercise=TRUE, exercise.setup="preLoadMap"}
my_storms <- storms %>%
  filter(year == 2016, name %in% c('Matthew', "Nicole", "Karl", "Otto"))

my_map + 
  coord_map('globular') + 
  geom_path(data = my_storms, aes(x = long, y = lat), col = 'red') +
  # geom_circle(
  #   data = my_storms, 
  #   aes(x0 = long, y0 = lat, r = tropicalstorm_force_diameter/120),
  #   col = 'red'
  # ) +
  coord_map('globular') +  # results in distorted storm circles
  guides(color = "none") + 
  theme_bw() + 
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  ) + 
  facet_wrap(~name) + 
  ggtitle("Selected 2016 Tropical Storms and Hurricanes")
```

The variable `tropicalstorm_force_diameter` gives the diameter in nautical miles of the area experiencing tropical storm force winds. 
Because 1 degree of latitude corresponds to approximately 60 nautical miles, we can approximately display the impact area of each storm by adding circles with radius `tropicalstorm_force_diameter/120`. 
Note that these circles can be distorted based on our projection. 
