---
title: "DATASCI 306: Lab 4"
subtitle: "Data Manipulation"
author: "Your Name"
output: 
  slidy_presentation:
    incremental: true
runtime: shiny_prerendered
---

```{r installPackages, echo=FALSE, include=FALSE, message=FALSE}
# This just checks if students need to install some packages that they might 
# not have.
if (!require(gapminder)) install.packages("gapminder", repos = "http://cran.us.r-project.org")

```

```{r setup, echo=FALSE, include=FALSE}
library(learnr)     # For interactive exercises
library(tidyverse)  # For ggplot2, dplyr
library(gapminder)
```

## Today's Content

> -   mutate, transmute, rename
> -   group_by, summarize
> -   pipe operator
> -   writing custom

## Mutate

We can use `mutate` to add new variables and preserve existing ones.
We will look at a few examples.
Here we use the dataset iris.

```{r iris}
head(iris)
```

## Basics of Mutate

> -   Create a Sepal Area column that equals to `Sepal Width * Sepal Length`

```{r}
iris1 <- mutate(iris, Sepal.Area = Sepal.Width * Sepal.Length)
head(iris1)
```

> -   Make this new column before Sepal Length

```{r}
iris1 <- mutate(iris, Sepal.Area = Sepal.Width * Sepal.Length,
                      .before = Sepal.Length)
head(iris1)
```

There are more useful arguments:

> -   `.keep = c("all", "used", "unused", "none")`
> -   `.before` and `.after` - optionally specifies the position of the new column

```{r mutate, exercise=T}
?mutate
```

## Creating Multiple Columns

You can create multiple columns, some of which may depend on those being created at the same time.

```{r}
## Create Sepal Area column as Sepal Width * Sepal Legth
## Update this area by scaling it by a factor of 10
iris1 <- mutate(iris, Sepal.Area = Sepal.Width * Sepal.Length,
               Scaled.Sepal.Area = Sepal.Area * 10)
head(iris1)
```

If we only want to keep the new columns and those columns used to create the new columns in the resulting dataset:

```{r}
iris1 <- mutate(iris, Sepal.Area = Sepal.Width * Sepal.Length,
               Scaled.Sepal.Area = Sepal.Area*10,
               .keep = "used")
head(iris1)
```

*Note*: We can use mutate to delete a column by setting the column name to NULL

```{r}
## Remove Species column
iris1 <- mutate(iris, Species = NULL)
head(iris1)
```

## Using Functions

> -   We can use function `row_number()` to get row id.
> -   Use function `toupper()` to convert characters to upper cases.

```{r}
iris1 <- mutate(iris, ID = row_number(),
                Species.Upper = toupper(Species))
head(iris1)
```

> -   Use `lag()` and `lead()` to get previous and later rows respectively.

```{r}
iris1 <- mutate(iris, previous = lag(Sepal.Length),
                      future = lead(Sepal.Length))
head(iris1)
```

> -   Use `case_when()` to create a conditional variable, `Size`: if `Sepal Length` is less than the average, assign the value "small", if it is greater than the average, assign "large", else assign "medium".

```{r}
iris1 <- mutate(iris, Size = case_when(
                 Sepal.Length < mean(Sepal.Length) ~ "small",
                 Sepal.Length > mean(Sepal.Length) ~ "large",
                   TRUE ~ "medium")
                   )
head(iris1)
```

\~ indicates the start of the formula - anonymous function - (which needs to return a value) to be executed if the preceding condition is `TRUE`, in this case a simple string.

Like an if statement, the arguments are evaluated in order, so you must proceed from the most specific to the most general.
For example, this won't work:

```{r}
x <- 1:50
case_when(
  TRUE ~ as.character(x),
  x %%  5 == 0 ~ "fizz",
  x %%  7 == 0 ~ "buzz",
  x %% 35 == 0 ~ "fizz buzz"
)
```

Let's try replicating the above with our own function:

```{r custom_function}
eval.sepal.length <- function(sepal_length, mean_length) {
  if (sepal_length < mean_length) {
    return("small")
  } else if (sepal_length > mean_length) {
    return("large")
  } else {
    return("medium")
  }
}
```

```{r}
  iris1 |>
  mutate(mean = mean(Sepal.Length)) |>
  rowwise() |> 
  mutate(Size = eval.sepal.length(Sepal.Length, mean))
```

We needed to use `rowwise` for this example since such simple functions can only be evaluated one row at the time, unlike `dplyr`'s `case_when`.

## Mutate Multiple Columns

```{r}
iris1 <- mutate(iris, across(starts_with("Sepal"), ~ . / 10)) 
head(iris1)
```

Breaking down the syntax above:

-   across: specifies a set of columns to mutate
-   \~ indicates the start of a formula to apply across several columns
-   . is a dummy variable (a *placeholder*) standing in for each column: it represents the argument passed to the nested `across` function by the outer `mutate` function (in this case).

## transmute

`transmute` works exactly like `mutate`, the only difference being that only newly created columns will be retained in our new dataset (tibble).
It is exactly the same as setting .keep = "none".

We can create Sepal Area and Scaled Sepal Area columns as a separate dataset

```{r}
iris2 <- transmute(iris, Sepal.Area = Sepal.Width * Sepal.Length,
               Scaled.Sepal.Area = Sepal.Area*10)
head(iris2)
```

Alternatively, we can create Sepal Area and Scaled Sepal Area columns as a separate dataset using mutate

```{r}
iris3 <- mutate(iris, Sepal.Area = Sepal.Width * Sepal.Length,
               Scaled.Sepal.Area = Sepal.Area*10,
               .keep = "none")
head(iris3)
```

We can check using `dplyr`'s `all.equal` function to compare the two datasets.

```{r}
all.equal(iris2, iris3)
```

## Rename

We use `rename` to rename existing column names.

dots(.) are not recommended for variable names, we want to rename all columns names in the iris dataset by replacing .
with \_

```{r}
iris2 <- rename(iris, sepal_length = Sepal.Length,
                      sepal_width = Sepal.Width,
                      petal_length = Petal.Length,
                      petal_width = Petal.Width)
head(iris2)
```

We have many repetitive codes above.
We can also use `rename_with()` to apply a (string) function to all columns by default or a specific subset.

```{r}
iris2 <- rename_with(iris, ~ str_replace(., pattern="\\.", replacement = "_"))
head(iris2)
```

> -   For all the columns in iris, we apply the function `str_replace()`.
> -   \~ again indicates the start of a formula which needs to return a calue
> -   the `.` inside `str_replace` is, again, a *placeholder* standing for whatever is passed to it by `rename_with`
> -   We use "\\\\" to escape the `.` since `str_replace` expects a regular expression by default, where `.` stands in for any character.

## Summarize and group_by

`summarize` is used to compute a new dataset (tibble) by using aggregation functions.
It is usually useful when used with `group_by`, as this helps to compute summaries for each category of a column of a dataset.

Some useful aggregation functions include:

> -   Center: mean(), median()
> -   Spread: sd(), IQR(), mad()
> -   Range: \`min(), max(), quantile()
> -   Position: first(), last(), nth()
> -   Count: n(), n_distinct()
> -   Logical: any(), all()

## Summarize for entire dataset

To find the mean of Sepal Width and Sepia Length for the iris dataset:

```{r}
summarise(iris, mean_sep_width = mean(Sepal.Width), 
          mean_sep_length = mean(Sepal.Length)
          )

```

Question: What does the code below do?

```{r}
summarise(iris, n = n(), 
             unique= n_distinct(Species),
             first = first(Species), 
             last = last(Species),
             `30th` = nth(Species, 30))
```

## Summarize by groups

We first group the iris dataset by Species.

```{r}
by_group <- group_by(iris, Species)

```

Then take the mean for Sepal Length and Width within each group:

```{r}
summarise(by_group, mean_sep_length = mean(Sepal.Length),
          mean_sep_width = mean(Sepal.Width))
```

## The Pipe Operator

As you probably saw from the previous example, it is extraneous to be storing intermediate steps while completing a chain of tasks.
The pipe operator (`|>`) helps up eliminate this hassle.
Note: An easier way to understand the pipe (%\>%) operator is reading it as "and then"

Consider how the pipe (`|>`) can help eliminate such hassle in the following sequence (chain) of tasks:

> 1.  Filter the iris dataset for Petal Length greater than 1 and Sepal Width less than 3.3
> 2.  Select Sepal Length, Sepal Width and Species
> 3.  Rename the columns of Sepal Width and Sepal Length by replacing dot(.) with underscore(\_)
> 4.  Convert the Species column to uppercase
> 5.  Group the resulting dataset by species and compute the 50th percentile and the mean for each group

```{r piping, exercise=TRUE}
```


## Exercises

We use the dataset `gapminder` for the exercises:

```{r}
head(gapminder)
```

It contains demographic/economic information (population, GDP per capita, life expectancy) for many of the countries in the world every five years, provided by the Gapminder project.

## Question 1

Complete the following tasks:

> -   subset for European and Asian countries excluding Japan, Norway and Israel
> -   scale the population by dividing by 1000
> -   select all columns aside from those ending with "y"
> -   sort the resulting tibble by continent ascending and gdp per capita in descending order
> -   Retain only rows 5 to 20

What is the first row in the resulting dataset?

```{r question1, exercise = T}

```

## Question 2

Complete the following tasks:

> -   group the dataset by year and continent
> -   compute the total population (hint: use sum()), mean life expectancy and count the number of cases for each group
> -   make a scatterplot of year by total population continent. The plots should be separate for each continent.

```{r question2, exercise = T}

```

## Question 3

-   subset for only the year 2007
-   group by continent
-   compute the mean life expectancy and count the number of cases for each group
-   create:
    -   a bar chart for continent, where the bars are horizontal than vertical
    -   create a pie chart for the counting variable
        -   fill by continent
        -   add texts superimposing the counts of each continent on the pie chart [hint]: geom_text or geom_label
        -   move the legend to the bottom position than the right position

```{r question3, exercise = T}
iris |>                                                ## then
    filter(Petal.Length > 1 & Sepal.Width < 3.3) |>    ## and then
    select(Sepal.Length, Sepal.Width, Species) |>      ## and then
    rename(sepal_length = Sepal.Length,
           sepal_width = Sepal.Width,) |>              ## and then
    mutate(Species = toupper(Species)) |>              ## and then
    group_by(Species) |>                               ## and then
    summarize(mean_sep_length = mean(sepal_length),
              mean_sep_width = mean(sepal_width),
              `50th_tile_sep_len` = quantile(sepal_length, 0.5),
              `50th_tile_sep_width` = quantile(sepal_width, 0.5)
              )
```


## Writing custom functions

### Predicates

Let's write a *predicate* function that evaluates whether the country's GDP is in the bottom quantile, with 20% as the default value:

```{r}
bottom.quantile <- function(x, q=0.2) {
  return(x < quantile(x, q))
}
```

Now use it to filter the gapminder dataset for countries with GDP in the bottom **10%**, after averaging the GDP over the years, and order in ascending order of averaged GDP:

```{r gpd-bottom-quantile, exercise=TRUE}
```


### Inline and anonymous functions

Now let's compute the GDP **growth rate** for a country and display the top 10 countries by their average GDP growth:

```{r gdpgrowth, exercise=TRUE}

```

Now, let's find all the (country, year) pairs that had a *negative* average *nominal GDP* growth, using `filter`, then count and display the occurrences of such per each country, starting with the country with the most:

*Tip: make sure you group in the correct order so you're not comparing GDP's across different countries.*

```{r negativegrowth, exercise=TRUE}

```


Using an *anonymous function*, let's ignore any rows with negative growth.

Where `~` is, again, the shorthand for anonymous functions:

```         
~ <=> function(x) { ... return something}
```

and `.` stands for the argument of the function.


*Tip: The above GDP growth is the only possible negative value in a row, after selecting the numeric columns:*

```{r negativegrowthanonymous, exercise=TRUE}

```


